---
title: "CONTENTO TESTING AND VALIDATION"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
---

## Setup

```{r}
# Load required libraries
library(here)
library(tidyverse)

# Source 
source(here("R/utils.R"))
```

Use this code to set up your project-specific .Renviron file. It requires the `usethis` package.

```{r, eval=FALSE}
install.packages("usethis") # Install if not already installed
usethis::edit_r_environ(scope = "project")
```

Add the path to your Box folder in the .Renviron file. For example: `BOX_PATH="/Users/isabelfe/Library/CloudStorage/Box-Box/KLEMONLAB_Projects/Shared_CONTENTO/testing_data"`

We will use data located in two locations:

- inst/extdata: Data included in the package for running of the vignettes. This data is included in the package and will be available to users who install the package.
- Shared_CONTENTO/testing_data: Data not included in the package. This data is used for testing and validation of the app.


```{r}
testdata_path <- Sys.getenv("BOX_PATH") 
extdata_path <- file.path(here("inst/extdata/"))


Human_annot <- readRDS(file.path(extdata_path, "genomes/Homo_sapiens.GRCh38.113_annot.rds"))
ExpSingleContrast_SE <- readRDS(file.path(extdata_path, "SE/ExpSingleContrast_SE.rds"))
```
