---
title: "Create SE for CONTENTO"
date: last-modified
execute:
  eval: false
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
---

# Overview

**CONTENTO** expects a single `SummarizedExperiment` (SE) object (`.rds`) containing expression data, differential expression results for every contrast, and descriptive metadata. This document walks through building that object using RSEM/tximport counts and DESeq2, but CONTENTO itself is agnostic to how the downstream analysis was performed. If you have a different quantification method (e.g. Salmon, STAR + featureCounts) or a different DE tool (e.g. edgeR, limma-voom), you can still build a compatible SE as long as you reproduce the following structure.

A CONTENTO-compatible SE has four required components:

- **assays**: must contain an assay named `expr_matrix`, which provides the expression values used by CONTENTO visualizations and downstream analyses.
- **rowData**: one row per gene, with DE statistics for each contrast stored as columns prefixed by statistic name and suffixed by contrast name (e.g. `log2FoldChange_VirusTvsF`, `padj_VirusTvsF`). CONTENTO parses these prefixes (`baseMean_`, `log2FoldChange_`, `log2FoldChange_shrunk_`, `lfcSE_`, `stat_`, `pvalue_`, `padj_`) to reshape the data internally.
- **colData**: sample-level metadata (experimental factors used for grouping/coloring in plots).
- **metadata(se)**: a list with, at minimum, `contrasts` (a character vector of unique contrast names matching the `rowData` suffixes described above). Optional fields `organism` and `annotation` unlock MSigDB-based GSEA and annotation loading, respectively (see the annotation article for details).

| Feature | Required SE component |
|------------------------------------|------------------------------------|
| DE tables & volcano plots | rowData contrast columns |
| Expression Plot / Heatmap / GESECA | `expr_matrix` assay |
| Variance Decomposition tab | optional `varpart_*` rowData columns |
| MSigDB GSEA | `metadata(se)$organism` matching a valid `msigdbr` species |
| Neighborhood Analysis (bacteria) | `metadata(se)$organism == "Bacteria"` + annotation with coordinates |

## Workflow

Both examples below follow the same sequence:

1.  **Metadata Preparation**: load sample metadata and set factor levels/labels.
2.  **Data Import**: pull RSEM quantifications into a `tximport` object.
3.  **Exploratory DESeq2 Analysis (Optional)**: fit a simple model and inspect a PCA to inspect the data before committing to a final design.
4.  **Differential Expression Analysis**: fit the full design and extract one or more contrasts.
5.  **Create Expression Matrix**: extract the expression matrix for use in CONTENTO (normalization recommended).
6.  **Create SummarizedExperiment (SE) Object**: assemble assays, rowData, colData, and metadata with `convert_to_SE()`.

**Example 1** covers a single two-group comparison (one contrast). **Example 2** covers a multi-factor design with an interaction term, producing several related contrasts and also includes variance partitioning results.

```{r}
# Set up your R environment
library(tidyverse)
library(here)
library(tximport)
library(DESeq2)
library(vsn)
library(edgeR)
library(variancePartition)
library(ashr)
library(SummarizedExperiment)

SE_DIR <- here("../inst/extdata/SE")
```

# Example 1: Single Contrast

A simple experiment examining RSV infection in HNO-ALI cultures across three independent experiments. Samples are classified by a single factor (`virus`), with levels `FALSE` (control) and `TRUE` (RSV infected).

## Step 1. Metadata Preparation

`meta` must contain one row per sample, a `file_base` column matching your RSEM output filenames, and the experimental factor columns used in the design formula below.

```{r}
# Read metadata file
project_name <- "ExpSingleContrast"
meta <- read_csv(file.path(SE_DIR, paste0(project_name, "_metadata.csv")))

# Create unique condition and label for each sample. This will be used by CONTENTO for visualization.
meta <- meta |>
  mutate(
    virus = factor(virus, levels = c("FALSE", "TRUE")),
    exp = factor(exp),
    label = interaction(virus, exp, sep = " | ") # unique label for each sample 
  )
```

## Step 2. Data Import

The [**tximport**](https://www.bioconductor.org/packages/release/bioc/html/tximport.html) [@tximport] function reads in gene-level quantification results from RSEM and organizes them into a structured list for downstream analysis. It extracts estimated counts, transcript lengths, and abundance values, ensuring compatibility with DESeq2 and other Bioconductor workflows. Since some transcripts may have a reported length of 0, we update those to 1 to prevent issues ensuring valid calculations during analysis.

```{r}
# Import data using tximport
rsem_folder <- file.path(SE_DIR, "rsem_counts")
count_files <- file.path(rsem_folder, paste0(meta$file_base, ".genes.results"))
names(count_files) <- meta$label
txi <- tximport(count_files, type = "rsem", txIn = FALSE, txOut = FALSE)
txi$length[txi$length == 0] <- 1
```

## Step 3. Exploratory DESeq2 Analysis

We run an initial DESeq model design to explore the data. This exploratory fit is only used for the PCA check below — it is not the design used for contrast extraction, which comes next.

```{r}
# DESeq design 
dds <- DESeqDataSetFromTximport(txi,
                                colData = meta,
                                design = ~ virus + exp)

# Run DESeq
dds <- DESeq(dds)
```

### Principal Component Analysis (PCA)

PCA calculated based on the VST normalized [@tibshirani1988; @huber2003; @Anders2010] data using ntop=500 top features by variance. Samples from different experiments were interspersed rather than clustered, suggesting that the experimental batch effect was not a major source of variation in the data.

```{r, fig.width=8, fig.height=6, fig.align="center"}
vsd <- vst(dds, blind = TRUE)
pcaData <- plotPCA(vsd, intgroup = c("virus", "exp"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
Bac.Vir_colors = c("#CC9564", "#856141")
Bac.Vir_fill <- c("white", "#856141")

ggplot(pcaData, aes(PC1, PC2)) +
  geom_point(aes(fill = virus, color = virus, 
                 shape = exp), 
             size = 4, stroke = 1) +  
  scale_shape_manual(values = c(23, 25, 21)) + 
  scale_color_manual(values = Bac.Vir_colors) + 
  scale_fill_manual(values = Bac.Vir_fill) +
  ggtitle("Principal Component Analysis (blind = TRUE)") +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
    guides(
    color = guide_legend(title = "virus", override.aes = list(color = Bac.Vir_colors, shape = 22))
  )
```

## Step 4. Differential Expression Analysis

The design captures all the variables of interest.

```{r}
# DESeq design 
dds <- DESeqDataSetFromTximport(txi,
                                colData = meta,
                                design = ~ virus + exp)

# Run DESeq
dds <- DESeq(dds)
```

### Generate Contrast Results

This function runs the DESeq2 `results()` and `lfcShrink()` functions for a given contrast or coefficient, returning a data frame with the results. It handles both types of contrasts (main effects and interaction terms) and adds the contrast name to the results for easier identification.

```{r}
run_contrast_analysis <- function(dds, contrast_name, contrast = NULL, coef = NULL) {
  if (!is.null(contrast)) {
    res <- results(dds, contrast = contrast, alpha = 0.05)
    resLFC <- lfcShrink(dds, contrast = contrast, type = "ashr")
  } else if (!is.null(coef)) {
    res <- results(dds, name = coef, alpha = 0.05)
    resLFC <- lfcShrink(dds, coef = coef, type = "ashr")
  } else {
    stop("Please provide either a 'contrast' or a 'coef'.")
  }

  res_df <- as.data.frame(res)
  res_df$Geneid <- rownames(res_df)
  res_df$contrast <- contrast_name

  res_df$log2FoldChange_shrunk <- as.data.frame(resLFC)$log2FoldChange[
    match(res_df$Geneid, rownames(resLFC))
  ]

  return(res_df)
}
```

```{r}
contrasts_df <- run_contrast_analysis(dds, contrast_name = "VirusTvsF", coef = "virus_TRUE_vs_FALSE")
```

## Step 5. Create Expression Matrix

The expression matrix stored in the SE can be generated using any normalization or transformation method appropriate for your dataset. CONTENTO treats this assay as a matrix of expression values used for visualization and downstream analyses such as expression plots, heatmaps, and GESECA. Common choices include VST-transformed counts from DESeq2, log-CPM values from edgeR, or log-transformed TPM values.

For RNA-seq workflows based on DESeq2, we recommend using the variance stabilizing transformation (VST), which reduces the dependence of variance on expression level while preserving biological signal. This generally produces expression values that are well suited for visualization and exploratory analyses in CONTENTO.

```{r}
expr_matrix <- assay(vst(dds, blind = FALSE))
```

## Step 6. Create SummarizedExperiment (SE) Object

The `convert_to_SE()` helper below builds the final SE object. It reshapes the long-format differential expression results into the wide-format `rowData` structure expected by CONTENTO, where each statistic is stored in a separate column with the pattern `<statistic>_<contrast_name>`. It attaches sample metadata as `colData`, stores the expression matrix in the `expr_matrix` assay, optionally merges variance-partition results as `varpart_*` columns, and records `contrasts`, `organism`, and `annotation` in `metadata(se)`. The same function can be reused regardless of experimental design or the number of contrasts being included.

```{r}
convert_to_SE <- function(dds, contrasts_df, expr_matrix,
                          organism = NULL, annotation = NULL, varpart = NULL) { 
  
  # Reshape to wide format
  contrasts_wide <- contrasts_df %>%
    pivot_wider(
      id_cols = Geneid,
      names_from = contrast,
      values_from = -c(Geneid, contrast),
      names_sep = "_"
    )
  
  contrasts_found <- unique(contrasts_df$contrast)
  cat("✓ Found", length(contrasts_found), "contrasts:", paste(contrasts_found, collapse = ", "), "\n")
  
  # Extract sample metadata
  sample_metadata <- as.data.frame(colData(dds))
  
  # Create rowData with Geneid as rownames
  rowdata_df <- contrasts_wide %>%
    column_to_rownames("Geneid")
  
  # Add variance partition data if provided
  if (!is.null(varpart)) {
    cat("Loading variance partition data...\n")
    
    tryCatch({
      # Extract only the variance partition matrix
      varpart_matrix <- varpart$varPart
      
      # Get common genes between rowData and varPart
      common_genes <- intersect(rownames(rowdata_df), rownames(varpart_matrix))
      
      if (length(common_genes) > 0) {
        # Convert to dataframe and add varpart_ prefix
        varpart_df <- as.data.frame(varpart_matrix[common_genes, , drop = FALSE])
        colnames(varpart_df) <- paste0("varpart_", colnames(varpart_df))
        
        # Merge with rowdata_df (only for genes that exist in both)
        rowdata_df[common_genes, colnames(varpart_df)] <- varpart_df
        
        cat("✓ Added variance partition for", length(common_genes), "genes\n")
        cat("  Factors:", paste(gsub("varpart_", "", colnames(varpart_df)), collapse = ", "), "\n")
      } else {
        cat("⚠ No common genes between contrasts and variance partition\n")
      }
      
    }, error = function(e) {
      cat("⚠ Error loading variance partition:", e$message, "\n")
    })
  }
  
  # Ensure rowData order matches dds
  rowdata_df <- rowdata_df[rownames(dds), ]
  
  # Verify dimensions match
  stopifnot(identical(rownames(rowdata_df), rownames(dds)))
  stopifnot(identical(rownames(sample_metadata), colnames(dds)))
  
  cat("✓ Gene order verified:", nrow(rowdata_df), "genes match\n")
  
  # Create SummarizedExperiment
  se <- SummarizedExperiment(
    assays = list(expr_matrix = expr_matrix),
    colData = DataFrame(sample_metadata),
    rowData = DataFrame(rowdata_df)
  )
  
  # Store metadata
  metadata(se) <- list(
    design = design(dds),
    contrasts = contrasts_found,
    organism = organism,
    annotation = annotation, 
    deseq2_version = as.character(packageVersion("DESeq2")),
    date = Sys.Date()
  )
  
  # Return the SummarizedExperiment object

  cat("  Dimensions:", nrow(se), "genes x", ncol(se), "samples\n")
  cat("  Assays:", paste(names(assays(se)), collapse = ", "), "\n")
  cat("  Columns in rowData:", ncol(rowData(se)), "\n\n")
  
  return(se)
}
```

```{r}
SE = convert_to_SE(
  dds = dds,
  contrasts_df = contrasts_df,
  expr_matrix = expr_matrix,
  organism = "Homo sapiens", 
  annotation = "Homo_sapiens.GRCh38.113_annot"
)
saveRDS(SE, file.path(SE_DIR, paste0(project_name, "_SE.rds")))
```

# Example 2: Multiple Contrasts

A more complex experiment examining RSV infection across four bacterial conditions (NB, Spn, Hin, and Dpi) in HNO-ALI cultures. The design models virus status, bacterial condition, and their interaction while accounting for variation across independent experiments.

## Step 1. Metadata Preparation

`meta` must contain one row per sample, a `file_base` column matching your RSEM output filenames, and the experimental factor columns used in the design formula below.

```{r}
# Read metadata file
project_name <- "ExpMultContrast"
meta <- read_csv(file.path(SE_DIR, paste0(project_name, "_metadata.csv")))

# Create unique condition and label for each sample. This will be used by CONTENTO for visualization.
meta <- meta |>
  mutate(
    bacteria = factor(bacteria, levels = c("NB", "Spn", "Hin", "Dpi")), # NB (No-Bacteria) is the reference level
    virus = factor(virus, levels = c("FALSE", "TRUE")),
    exp = factor(exp),
    condition = interaction(bacteria, virus, sep = "_"), # unique condition for each sample
    label = interaction(bacteria, virus, exp, sep = " | ") # unique label for each sample 
  )
```

## Step 2. Data Import

Same import procedure as Example 1, applied to this project's samples.

```{r}
# Import data using tximport
rsem_folder <- file.path(SE_DIR, "rsem_counts")
count_files <- file.path(rsem_folder, paste0(meta$file_base, ".genes.results"))
names(count_files) <- meta$label
txi <- tximport(count_files, type = "rsem", txIn = FALSE, txOut = FALSE)
txi$length[txi$length == 0] <- 1
```

## Step 3. Exploratory DESeq2 Analysis

We run an initial DESeq model design to explore the data. It does not consider the interaction between variables. We only use it to explore the data and possible effects for the 3 independent experiments. The design formula is `~ bacteria + virus + exp`, which accounts for the main effects of bacteria, virus, and experiment without considering their interactions. This allows us to assess the individual contributions of each factor to gene expression changes. This exploratory fit is only used for the PCA check below, it is not the design used for contrast extraction, which comes later.

```{r}
# DESeq design 
dds <- DESeqDataSetFromTximport(txi,
                                colData = meta,
                                design = ~ virus + bacteria + exp)

# Run DESeq
dds <- DESeq(dds)
```

### Principal Component Analysis (PCA)

PCA calculated based on the VST normalized [@tibshirani1988; @huber2003; @Anders2010] data using ntop=500 top features by variance. Samples from different experiments were interspersed rather than clustered, suggesting that the experimental batch effect was not a major source of variation in the data.

```{r, fig.width=8, fig.height=6, fig.align="center"}
vsd <- vst(dds, blind = TRUE)
pcaData <- plotPCA(vsd, intgroup = c("bacteria", "virus", "exp"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
Bac.Vir_colors = c("#CC9564", "#a89de0", "#f54590", "#2e67f2", "#856141","#6855c9","#9f2d5e", "#1e439d")
Bac.Vir_fill <- c("white", "white", "white", "white", "#856141","#6855c9","#9f2d5e", "#1e439d")

ggplot(pcaData, aes(PC1, PC2)) +
  geom_point(aes(fill = condition, color = condition, 
                 shape = exp), 
             size = 4, stroke = 1) +  
  scale_shape_manual(values = c(23, 25, 21)) + 
  scale_color_manual(values = Bac.Vir_colors) + 
  scale_fill_manual(values = Bac.Vir_fill) +
  ggtitle("Principal Component Analysis (blind = TRUE)") +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
    guides(
    color = guide_legend(title = "condition", override.aes = list(color = Bac.Vir_colors, shape = 22))
  )
```

### Variance Partitioning Analysis

In this more complex design we are also interested in quantifying the contribution of each factor (virus, bacteria, and experiment) to the overall variance in gene expression. This is particularly important for understanding how much of the observed variability can be attributed to biological factors versus technical or experimental factors. CONTENTO allows users to visualize the variance partitioning results for each gene, providing insights into the relative importance of each factor in shaping gene expression patterns.

Gene-level expression values were modeled using precision weights estimated with the `limma::voom()` transformation, which accounts for the mean–variance relationship in RNA-seq data. Lowly expressed genes can be filtered using `edgeR::filterByExpr()`, which removes genes with insufficient counts for reliable statistical analysis and improves the stability of downstream variance estimates.

We use the [variancePartition](https://bioconductor.org/packages/release/bioc/html/variancePartition.html) [@hoffman2017] package, which quantifies the proportion of variance in gene expression attributable to each factor in a linear mixed model.

```{r}
run_varPart_analysis <- function(txi, meta, formula_vp, formula_fixed, filter_low_expr = TRUE) {
  # Align meta to sample order
  meta <- as.data.frame(meta)
  meta <- meta[match(colnames(txi$counts), meta$label), ]
  rownames(meta) <- meta$label
  stopifnot(identical(rownames(meta), colnames(txi$counts)))

  # Design matrix: only fixed effects (for voom & filterByExpr)
  design <- model.matrix(formula_fixed, data = meta)

  # Normalize counts
  counts <- DGEList(txi$counts)
  counts <- normLibSizes(counts)

  # Filter low-expressed genes
  if (filter_low_expr) {
    keep <- filterByExpr(counts, design)
    counts_used <- counts[keep, ]
  } else {
    counts_used <- counts
  }

  # logCPM values with voom
  v <- voom(counts_used, design)

  # Fit variance partition model (supports random effects)
  varPart <- fitExtractVarPartModel(v, formula_vp, meta)

  # Format results
  varPart_df <- as.data.frame(varPart)
  varPart_df$Geneid <- rownames(varPart_df)

  # Return
  return(list(v = v, varPart = varPart, varPart_df = varPart_df))
}
```

Use formula_fixed with fixed effects (required for the design matrix step) while formula_vp is used to decompose variance across bacteria/virus/exp as random intercepts

```{r, fig.width=8, fig.height=5, fig.align="center"}
formula_fixed <- ~ virus + bacteria + exp          # for voom/filterByExpr design matrix
formula_vp <- ~ (1|virus) + (1|bacteria) + (1|exp)   # for variance partitioning you can't mix fixed and random categorical variables in the same call 

varPart <- run_varPart_analysis(txi, meta, formula_vp, formula_fixed, filter_low_expr = T)
plotVarPart(varPart$varPart) 
```

## Step 4. Differential Expression Analysis

This design captures all the variables of interest. - **virus + bacteria + virus:bacteria** Includes main effects of virus and bacteria, as well as their interaction. This allows us to test whether the effect of virus depends on the species. This is equivalent to **virus \* bacteria**. - **exp** To account for the batch effect of the 3 independent experiments, we include the experiment as a covariate in the model. This helps to control for any variability introduced by differences between experiments.

```{r}
# DESeq design 
dds <- DESeqDataSetFromTximport(txi,
                                colData = meta,
                                design = ~ virus + bacteria + virus:bacteria + exp)

# Run DESeq
dds <- DESeq(dds)
```

### Generate Contrast Results

#### Virus TRUE vs FALSE in each Species

This section tests the effect of virus infection (TRUE vs FALSE) within each bacterial species specifically:

- For the reference bacteria condition (NB), the virus effect is captured by the main coefficient `virus_TRUE_vs_FALSE`.
- For other bacteria conditions (Spn, Hin, Dpi), the virus effect is a combination of the main virus effect plus the interaction term (e.g., `virusTRUE.bacteriaDpi`).

```{r}
contrast_list_Virus_TvsF <- list(
  VirusTvsF_NB  = "virus_TRUE_vs_FALSE",                                   
  VirusTvsF_Spn = list(c("virus_TRUE_vs_FALSE", "virusTRUE.bacteriaSpn")), 
  VirusTvsF_Hin = list(c("virus_TRUE_vs_FALSE", "virusTRUE.bacteriaHin")), 
  VirusTvsF_Dpi = list(c("virus_TRUE_vs_FALSE", "virusTRUE.bacteriaDpi"))  
)
results_Virus_TvsF <- imap(contrast_list_Virus_TvsF, function(contrast_val, name) {
  if (is.character(contrast_val)) {
    run_contrast_analysis(dds, contrast_name = name, coef = contrast_val)
  } else {
    run_contrast_analysis(dds, contrast_name = name, contrast = contrast_val)
  }
})
```

#### Difference in Virus Effect Compared to NB

Here we focus on whether the virus effect differs between each bacteria condition and relative to the reference NB. This highlights species-specific modulation of virus response by comparing the virus effect in each bacteria against the baseline virus effect in NB.

- We use the interaction terms alone (e.g., `virusTRUE.bacteriaDpi`), which quantify how much the virus effect in each species deviates from that in NB.

```{r}
contrast_list_VirusInteract_vsNB <- list(
  VirusInteract_Dpi = "virusTRUE.bacteriaDpi",
  VirusInteract_Spn = "virusTRUE.bacteriaSpn",
  VirusInteract_Hin = "virusTRUE.bacteriaHin"
)
results_VirusInteract_vsNB <- imap(contrast_list_VirusInteract_vsNB, ~ run_contrast_analysis(dds, contrast_name = .y, coef = .x,))
```

#### Bacteria vs NB in Virus FALSE

This comparison evaluates how each bacterial species differs from the reference NB when virus is absent (`virus = FALSE`). This reveals baseline species-specific gene expression differences without virus influence.

- Because virus FALSE is the baseline in the model, these contrasts use only the main bacteria effects (e.g., `bacteria_Dpi_vs_NB`).

```{r}
contrast_list_Bacteria_vsNB_VF <- list(
  DpivsNB_VF = "bacteria_Dpi_vs_NB",
  SpnvsNB_VF = "bacteria_Spn_vs_NB",
  HinvsNB_VF = "bacteria_Hin_vs_NB"
)
results_Species_vsNB_VF <- imap(contrast_list_Bacteria_vsNB_VF, ~ run_contrast_analysis(dds, contrast_name = .y, coef = .x))
```

#### Bacteria vs NB in Virus TRUE

Here we test the bacteria differences under virus infection (virus = TRUE). This reveals species-specific gene expression differences that might occur under virus influence.

- These contrasts combine the main bacteria effect plus the interaction with virus (e.g., `bacteria_Dpi_vs_NB` + `virusTRUE.bacteriaDpi`).

```{r}
contrast_list_Bacteria_vsNB_VT <- list(
  DpivsNB_VT = list(c("bacteria_Dpi_vs_NB", "virusTRUE.bacteriaDpi")),
  SpnvsNB_VT = list(c("bacteria_Spn_vs_NB", "virusTRUE.bacteriaSpn")),
  HinvsNB_VT = list(c("bacteria_Hin_vs_NB", "virusTRUE.bacteriaHin"))
)
results_Bacteria_vsNB_VT <- imap(contrast_list_Bacteria_vsNB_VT, ~ run_contrast_analysis(dds, contrast_name = .y, contrast = .x))
```

#### Combine All Contrast Results

```{r}
all_contrast_results <- list(
  results_Virus_TvsF,
  results_VirusInteract_vsNB,
  results_Species_vsNB_VF,
  results_Bacteria_vsNB_VT
  ) %>% flatten() 

all_contrast_df <- bind_rows(all_contrast_results)
```

## Step 5. Create Expression Matrix

As in Example 1, the expression matrix can be generated using any normalization or transformation method suitable for downstream visualization and exploratory analysis. Here we use the DESeq2 variance stabilizing transformation (VST).

```{r}
expr_matrix <- assay(vst(dds, blind = FALSE))
```

## Step 6. Create SummarizedExperiment (SE) Object

We use the same `convert_to_SE()` function as in Example 1 to build the final SE object. In this case, we also include the variance partitioning results as additional columns in `rowData`, which can be visualized in CONTENTO's Variance Decomposition tab.

```{r}
SE = convert_to_SE(
  dds = dds,
  contrasts_df = all_contrast_df,
  expr_matrix = expr_matrix,
  varpart = varPart,
  organism = "Homo sapiens", 
  annotation = "Homo_sapiens.GRCh38.113_annot"
)
saveRDS(SE, file.path(SE_DIR, paste0(project_name, "_SE.rds")))
```

# Verify Your SE Object

Before uploading to CONTENTO, confirm:

- [ ] `assays(se)` includes `expr_matrix` (required for expression/heatmap/GESECA features).
- [ ] `rowData(se)` has `<stat>_<contrast_name>` columns for every entry in `metadata(se)$contrasts`, with matching suffixes.
- [ ] `metadata(se)$contrasts` is a character vector, not a factor or list.
- [ ] `metadata(se)$organism`, if set, exactly matches a valid `msigdbr` species name (or is `"Bacteria"` for neighborhood analysis).
- [ ] `metadata(se)$annotation`, if set, matches the filename (without `.rds`) of a built-in or uploaded annotation file.
- [ ] `rownames(se)` are consistent gene identifiers matching your annotation's `Geneid` column, if used.

If a tab or control is missing/greyed out in CONTENTO, it's almost always one of the above.
