---
title: "Introduction"
date: last-modified
execute:
  eval: false
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
---


# CONTENTO
> _CONTrast ExploratioN Toolkit for Omics_

## Intro

Shiny App for RNASeq Results Exploration.

## Running CONTENTO

CONTENTO can be executed in three different ways depending on your computational environment and familiarity with containerized software. We strongly recommend the Docker-based workflows, as they guarantee complete reproducibility across operating systems and avoid dependency conflicts.

### _Option 1:_ Run using Docker (recommended)

If Docker is not yet installed on your system, we recommend you install Docker Desktop from the [official website](https://www.docker.com/products/docker-desktop/), since it provides a friendly graphic interface to manage your images and keep your Docker Engine up-to-date. If you already have Docker installed, running CONTENTO requires a single command from your terminal _(make sure your Docker Desktop is running)_:

```{bash}
docker run \
    --rm \
    -p 3838:3838 \
    -v $(pwd)/inst:/srv/shiny-server/app/inst \
    ghcr.io/klemonlab/contento:latest
```

The terminal window will show the app startup and soon will start accepting connections. When you see the message that the app is listening, open your web browser (any browser) and navigate to:

```
http://localhost:3838
```

This approach provides a fully reproducible computational environment with all R, Bioconductor, and system dependencies pre-installed.



> **Rebuilding image from scratch**
>
> A Dockerfile is included in the repository so that users familiar with building container images
> can self-build with `docker build -t contento .` if necessary.

### _Option 2:_ Run using Docker Compose

For users who prefer a persistent local setup and the convenience of a preconfigured container mount, **CONTENTO** can also be launched using [Docker Compose](https://docs.docker.com/compose/). To do this, make sure you have the `compose` plugin installed (it is bundled with modern distributions of Docker Desktop by default) by running

```{bash}
docker compose --help
```
The helper string for the `compose` plugin should come up. 


Then, clone the CONTENTO repository:

```{bash}
git clone https://github.com/klemonlab/contento.git
```

To start the application, navigate to the working directory and simply run:

```{bash}
cd contento
docker compose up
```

Or run it in the background:

```{bash}
docker compose up -d
```

Once the container starts, access the application again at:

```
http://localhost:3838
```

### _Option 3:_ Run locally within R/RStudio

Some users may prefer to execute **CONTENTO** directly from an existing R installation, usually used in conjunction with the popular IDE [RStudio](https://posit.co/download/rstudio-desktop).

Clone the CONTENTO repository:

```{bash}
git clone https://github.com/klemonlab/contento.git
```

After opening the project (`CONTENTO.Rproj`) within Rstudio, install the dependencies by running:

```{r}
install.packages(c(
    "shiny",
    "shinydashboard",
    "shinycssloaders",
    "DT",
    "plotly",
    "tidyverse",
    "purrr",
    "viridis",
    "RColorBrewer",
    "ComplexUpset",
    "scales",
    "ggiraph",
    "gridExtra",
    "patchwork"
))

if (!requireNamespace("BiocManager"))
    install.packages("BiocManager")

BiocManager::install(c(
    "ComplexHeatmap",
    "DESeq2",
    "fgsea",
    "msigdbr",
    "circlize",
    "SummarizedExperiment"
))
```


Open `app.R` in RStudio and click the interactive  **:arrow_forward: Run App** button, or execute:

```{r}
shiny::runApp("app.R")
```

The application will then become available in your web browser.

---

### Which option should I choose?

| Method         | Recommended for               | Widely Reproducible    | Requires installation    |
| -------------- | ----------------------------- | ---------------------- | ------------------------ |
| Docker         | Most users                    | ✓                      | Docker                   |
| Docker Compose | Persistent local deployments  | ✓                      | Docker + Compose         |
| R/RStudio      | RStudio users and developers  | Depends on environment | R + package installation |

## Citation

If you use this application in scientific work, please cite:

Isabel FE, et al. "CONTENTO: CONTrast ExploratioN Toolkit for Omics". 
https://github.com/KLemonLab/CONTENTO
DOI: [add Zenodo DOI once available].

## License

This project is covered under the GPL-3 license.

## Contact

Issues, bug reports, and feature requests are welcome through the GitHub issue tracker.