---
title: "Create SE for CONTENTO"
date: last-modified
execute:
  eval: false
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
---

# Overview

CONTENTO expects a single `SummarizedExperiment` (SE) object, saved as `.rds`,
containing your DESeq2 (or equivalent) contrast results, a variance-stabilized
expression matrix, and a small set of metadata fields that drive routing
throughout the app (organism detection, MSigDB eligibility, gene labels,
download filenames, etc.).

This document describes, piece by piece, how to build that object from a
typical DESeq2 workflow with multiple contrasts. Adapt the `dds`/`res` object
names to your own pipeline — the important part is the **shape** of the final
SE, not the tool used to generate it.

::: callout-important
### Minimum requirements

- `assay(se, "vst")` — variance-stabilized expression matrix (genes x samples)
- `rowData(se)` — per-contrast columns named `<statistic>_<contrast>`
- `metadata(se)$contrasts` — character vector of contrast names, matching the
  suffixes used in `rowData(se)`
- `rownames(se)` — gene identifiers (Ensembl IDs, locus tags, etc.), matching
  the annotation file's `Geneid` column if you use one
:::

------------------------------------------------------------------------

# Step 1: Build the rowData contrast columns

For each contrast, CONTENTO looks for columns following the pattern
`baseMean_<contrast>`, `log2FoldChange_<contrast>`,
`log2FoldChange_shrunk_<contrast>`, `lfcSE_<contrast>`, `stat_<contrast>`,
`pvalue_<contrast>`, and `padj_<contrast>` (see `extract_de_results()` in
`utils.R`). Any of these can be missing for a given contrast — missing columns
are filled with `NA` — but `stat_<contrast>` is required if you want to run
GSEA for that contrast (used as the ranking statistic).

```{r}
library(DESeq2)
library(tidyverse)

# Example: dds already fit with DESeq(); contrasts is a named list of
# contrast specifications passed to results()/lfcShrink()
contrasts_list <- list(
  "treated_vs_control"   = c("condition", "treated", "control"),
  "mutant_vs_wildtype"   = c("genotype", "mutant", "wildtype")
)

rd <- data.frame(row.names = rownames(dds))

for (cname in names(contrasts_list)) {
  res        <- results(dds, contrast = contrasts_list[[cname]])
  res_shrunk <- lfcShrink(dds, contrast = contrasts_list[[cname]], type = "ashr")

  rd[[paste0("baseMean_", cname)]]             <- res$baseMean
  rd[[paste0("log2FoldChange_", cname)]]        <- res$log2FoldChange
  rd[[paste0("log2FoldChange_shrunk_", cname)]] <- res_shrunk$log2FoldChange
  rd[[paste0("lfcSE_", cname)]]                  <- res$lfcSE
  rd[[paste0("stat_", cname)]]                   <- res$stat
  rd[[paste0("pvalue_", cname)]]                 <- res$pvalue
  rd[[paste0("padj_", cname)]]                   <- res$padj
}
```

------------------------------------------------------------------------

# Step 2: Add the VST assay

The `"vst"` assay is used for the expression plots, heatmaps, and GESECA
co-expression analysis. It must be named exactly `"vst"`.

```{r}
vst_mat <- assay(vst(dds, blind = FALSE))
```

------------------------------------------------------------------------

# Step 3: Assemble the SE object

```{r}
se <- SummarizedExperiment(
  assays  = list(vst = vst_mat),
  rowData = DataFrame(rd),
  colData = colData(dds)
)
```

------------------------------------------------------------------------

# Step 4: Set metadata fields

All of these live in `metadata(se)` and are optional except `contrasts`,
which is required for the app to load at all (`extract_de_results()` errors
without it).

```{r}
metadata(se)$contrasts <- names(contrasts_list)

# Organism name — must exactly match a msigdbr species name (e.g.
# "Homo sapiens", "Mus musculus") to enable MSigDB gene sets, or the literal
# string "Bacteria" to enable the Genomic Neighborhood Analysis tab.
metadata(se)$organism <- "Homo sapiens"

# Name of a built-in annotation .rds file (without extension) shipped in
# inst/annotations/ — see create-genome-annotations.qmd. Leave NULL/omit to
# require the user to upload their own annotation file instead.
metadata(se)$annotation <- "Homo_sapiens.GRCh38.113"

# Optional: force MSigDB db_species ("HS" or "MM") instead of deriving it
# from `organism`, e.g. for a species only supported via ortholog mapping.
metadata(se)$db_species <- NULL

# Optional: used only in the GESECA condition-selector default
# (conditionSelectUI_geseca in compare_contrast_server.R).
metadata(se)$design_formula <- "~ condition"

# Optional: used as the base name for all download filenames
# (build_download_filename() in utils.R). Falls back to the uploaded
# file name if omitted.
metadata(se)$dataset_name <- "MyStudy_RNAseq"
```

------------------------------------------------------------------------

# Step 5 (Optional): Add variance partition data

If you've run `variancePartition::fitExtractVarPartModel()`, store the result
under `varpart_` prefixed columns in `rowData(se)` — `load_files_server.R`
detects any column matching `^varpart_` and builds the Variance Decomposition
tab automatically.

```{r}
library(variancePartition)

varPart <- fitExtractVarPartModel(vst_mat, ~ (1|condition) + (1|batch), colData(dds))

for (factor_name in colnames(varPart)) {
  rowData(se)[[paste0("varpart_", factor_name)]] <- varPart[[factor_name]]
}
```

------------------------------------------------------------------------

# Step 6: Save the SE object

```{r}
saveRDS(se, "MyStudy_SE.rds")
```

------------------------------------------------------------------------

# Step 7: Verify your SE file

Before uploading, confirm:

- [ ] `assay(se, "vst")` exists and has the same genes/samples as `rowData`/`colData`.
- [ ] `metadata(se)$contrasts` is a character vector, and each name has a
  matching set of `<statistic>_<contrast>` columns in `rowData(se)`.
- [ ] `rownames(se)` match the `Geneid` values in your annotation file exactly
  (same case, same format).
- [ ] If using MSigDB: `metadata(se)$organism` exactly matches a `msigdbr`
  species name, and either `rownames(se)` are Ensembl IDs or your annotation
  file has a detectable Ensembl ID column.
- [ ] If using the Bacteria neighborhood tab: `metadata(se)$organism` is
  exactly `"Bacteria"` and the annotation file has `start`, `end`, `strand`.
- [ ] File is saved as `.rds` and contains a `SummarizedExperiment` object
  (not a `DESeqDataSet` or list).

If a feature is missing in the app (e.g. no contrasts show up, MSigDB radio
absent, Variance Decomposition tab missing), it is almost always one of the
above checks failing silently — see `create-genome-annotations.qmd` for the
matching annotation-file checklist.
