---
title: "Create Bacterial Genome Annotations"
author: "Isabel Fernandez Escapa"
date: last-modified
execute:
  eval: false
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
---

# Overview

This tutorial shows how to create a functional annotation CSV file with gene IDs, descriptions, and annotations to be uploaded into the CONTENTO App.

------------------------------------------------------------------------

# Step 1: Source Genome Annotations

## 1.1 Get Your Genome from NCBI

1.  Go to [NCBI Assembly](https://www.ncbi.nlm.nih.gov/assembly/)
2.  Search for your genome accession (e.g., `GCF_000013465.1` for *Staphylococcus aureus* subsp. aureus FPR3757)
3.  Click on the result and downoload the assembly in GFF3 format.
4.  Unzip the file and save it in your working directory.

**Alternative: Use NCBI Datasets CLI**

```{bash, eval = FALSE}
# Install NCBI Datasets CLI if you haven't already. This should be run in your terminal.
conda activate ncbi_datasets 

# Update the genome accession and directory as needed
FILES_DIR="inst/extdata/genomes" 
GENOME_NAME="GCF_000013465.1" 

# Update to download the desired files (genome sequence, GenBank flat file, GFF3...)
datasets download genome accession "$GENOME_NAME" \
            --include genome,gbff,gff3,seq-report \
            --filename "$FILES_DIR/$GENOME_NAME.zip"
            
# Unzip
unzip "$FILES_DIR/$GENOME_NAME.zip" -d "$FILES_DIR"
# Individual files can be renamed to include the genome name for easier identification.
```

## 1.2 Generate your own GFF3 file (Optional)

For genomes not available in NCBI, you can generate your own GFF3 file using Prokka [@Seemann2014] or similar annotation tools. Just make sure to include the necessary attributes (e.g., `locus_tag`, `product`, ...) in the output GFF3.

------------------------------------------------------------------------

# Step 2: Parse Annotation into CSV

CONTENTO requires a CSV file with gene IDs, descriptions, and annotations. The GFF3 file contains this information in the `attributes` column, which we can parse to create a structured annotation CSV.

```{r}
# Set up your R environment and read the GFF3 file. Adapt based on the GFF3 name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

gff_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, ".gff"))

# Read the GFF3 file into a data frame
gff_df <- read_delim(
  gff_file,
  delim = "\t",
  col_names = c("seqname", "source", "feature", "start", "end", "score", "strand", "frame", "attributes"),
  comment = "#",
  show_col_types = FALSE
)

# Filter for CDS (protein-coding genes)
gff_cds <- gff_df |>
  filter(feature == "CDS")

# Function to parse the attributes column of GFF3
parse_gff3_attributes <- function(attributes_string) {
  pairs <- str_split(attributes_string, ";")[[1]]
  
  attr_list <- list()
  
  for (pair in pairs) {
    if (pair == "") next
    
    parts <- str_split(pair, "=", n = 2)[[1]]
    
    if (length(parts) == 2) {
      key <- parts[1]
      value <- parts[2]
      attr_list[[key]] <- value
    }
  }
  
  return(attr_list)
}

# Apply the parsing function to the attributes column and create new columns for each key
annotations <- gff_cds |>
  rowwise() |>
  mutate(attr_list = list(parse_gff3_attributes(attributes)))  |>
  ungroup() |>
  # Unnest all key=value pairs as columns
  unnest_wider(attr_list, names_repair = "universal")

# Save the annotations to a CSV file
output_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.csv"))
write_csv(annotations, output_file)
```

# Step 3: Add Additional Annotations (Optional)

The GFF3 file provides basic gene information, but you can significantly enhance your annotations by running additional bioinformatic tools.

These annotations are useful in CONTENTO for:

- **Functional enrichment analysis (GSEA)** using for example COG and KEGG categories
- **Gene exploration** - displaying comprehensive functional information for individual genes

## 3.1 Merge Annotations with the GFF3 derived annotations

```{r}
# Set up your R environment and read the anvi'o annotation file. Adapt based on the file name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")

# Read the custom annotation file (e.g., from RegPrecise regulon database or other source)
custom_file <- read_csv(file.path(GENOMES_DIR, "USA300_FPR3757_regulons.csv"))
```

## 3.2 Enhance Annotations with Anvi'o

**Anvi'o** is a comprehensive platform that integrates multiple annotation sources:

- **HMMs** - Generic protein domain detection
- **COG24** - Clusters of Orthologous Groups (excellent for functional enrichment)
- **KEGG** - Kyoto Encyclopedia of Genes and Genomes pathways (excellent for functional enrichment)
- **Pfams** - Protein family annotations
- **CAZymes** - Carbohydrate-active enzyme classification

```{bash, eval = FALSE}
# Install Anvio if you haven't already. This should be run in your terminal.
conda activate anvio-9

# Update the genome accession and directory as needed
FILES_DIR="inst/extdata/genomes" 
GENOME_NAME="GCF_000013465.1" 

# Process the GenBank file to extract gene calls, functions, and fasta sequences 
anvi-script-process-genbank -i "$FILES_DIR/$GENOME_NAME.gbff" \
--output-gene-calls "$FILES_DIR/${GENOME_NAME}_calls.txt" \
--output-functions "$FILES_DIR/${GENOME_NAME}_prodigal.txt" \
--output-fasta "$FILES_DIR/${GENOME_NAME}.fa" \
--annotation-source prodigal \
--include-locus-tags-as-functions

# Create anvi'o contigs database from the fasta file and gene calls
anvi-gen-contigs-database -f "$FILES_DIR/${GENOME_NAME}.fa" \
-o "$FILES_DIR/${GENOME_NAME}_anvio.db" \
--external-gene-calls "$FILES_DIR/${GENOME_NAME}_calls.txt" \
-n "$GENOME_NAME"

# Import functions from prodigal output
anvi-import-functions -c "$FILES_DIR/${GENOME_NAME}_anvio.db" \
-i "$FILES_DIR/${GENOME_NAME}_prodigal.txt"

# Run HMM annotations 
anvi-run-hmms -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run NCBI COG annotations with DIAMOND
anvi-run-ncbi-cogs -T 8 --search-with diamond -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run KEGG KOfam annotations
anvi-run-kegg-kofams -T 8 -E 1e-15 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run Pfam domain annotations
anvi-run-pfams -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run CAZyme annotations
anvi-run-cazymes -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Export all functions to text file
anvi-export-functions -c "$FILES_DIR/${GENOME_NAME}_anvio.db" \
-o "$FILES_DIR/${GENOME_NAME}_annot.txt"

# Export contigs stats and database info
anvi-display-contigs-stats "$FILES_DIR/${GENOME_NAME}_anvio.db" \
--report-as-text -o "$FILES_DIR/${GENOME_NAME}_stats.txt"

anvi-db-info "$FILES_DIR/${GENOME_NAME}_anvio.db" &>> "$FILES_DIR/${GENOME_NAME}_log.txt"
```

The resulting anvi'o annotation file contains comprehensive functional annotations for each gene, which can be parsed into a CSV format with the following R code.

```{r}
# Set up your R environment and read the anvi'o annotation file. Adapt based on the file name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

# Read the anvi'o annotation 
anvio_annot <- read_delim(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.txt")), delim = "\t") |>
  mutate(source = as.factor(source)) |>
  mutate(across(where(is.character), trimws)) |>
  select(-e_value) |> # Exclude e-values to keep dataframe simple
  rename("acc" = "accession", "func" = "function") # Shorter names

# Generate wide dataframe collapsing duplicates for the same source
anvio_annot <- anvio_annot |>
  pivot_wider(
    names_from = source,
    values_from = c(acc, func),
    values_fn = list(acc = ~ paste(unique(.), collapse = ","), func = ~ paste(unique(.), collapse = ","))
  ) 

# Sanity check for duplicates by GENBANK_LOCUS_TAG. It should be 0!!
nrow(anvio_annot |> group_by(acc_GENBANK_LOCUS_TAG) |> filter(n() > 1))
```

You can also merge the anvi'o annotations with the aa sequences from the calls file to have a comprehensive annotation dataframe with all the information for each gene. This can be useful for gene exploration in CONTENTO, where you can display the sequence along with the functional annotations for each gene.

```{r}
# Read calls file to get aa sequences
anvio_calls <- read_delim(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_calls.txt")), delim = "\t") |>
  select(gene_callers_id, aa_sequence)

# Merge with anvio_annot by gene_callers_id
anvio_annot <- anvio_annot |>
  left_join(anvio_calls, by = "gene_callers_id")
```
