---
title: "Create Genome Annotations for CONTENTO"
date: last-modified
execute:
  eval: false
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
---

# Overview:

Annotation files are optional, but they unlock several features in CONTENTO. Depending on your organism and the information included on your SummarizedExperiment (SE) object, your experience with CONTENTO can be highly limited without an annotation file, specifically for functional enrichment analysis and gene labeling.

## Annotation files for functional enrichment analysis

CONTENTO can run functional enrichment analysis (GSEA/GESECA) using gene set collections from two sources:

1.  **Annotation**: Available for any organism via user-provided data frame (in `.rds` format) with the following:

    - A `Geneid` column that matches the `rownames()` of your SE object.
    - At least one functional column (e.g. COG categories, KEGG pathways, custom gene sets, etc.) to be used as source for enrichment. Multiple values per gene should be separated by `!!!` (e.g., `"GO:0006415!!!GO:0016149"`) to be parsed correctly by counting the gene as belonging to each category for functional enrichment analysis.

2.  [**MSigDB**](https://www.gsea-msigdb.org/gsea/msigdb): Available only for human, mouse, and other species supported by [msigdbr](https://igordot.github.io/msigdbr/reference/msigdbr_species.html) R package. It requires:

    - A `metadata(se)$organism` that matches the `msigdbr` species name exactly (e.g. `Homo sapiens`, `Mus musculus`, etc.).
    - SE's `rownames()` should be Ensembl-style gene IDs (`ENSG...`, `ENSMUSG...`, etc.) or a column in the annotation file that contains Ensembl-style IDs. CONTENTO will auto-detect this column.
    

## Other benefits of annotation files

Annotation files also enable:

- **Gene symbols/labels** in place of raw Geneids throughout tables and plots. This is especially useful for bacteria, where `Geneid` is often a locus tag (e.g. `b0001`) and not not an informative label.
- **Genomic neighborhood analysis**, which additionally requires `start`, `end`, `strand` columns in the annotation file. This is a beta feature available only when `metadata(se)$organism` is exactly "Bacteria")

## Examples: Do I need an annotation file?

| SE's organism name | SE rownames | Desired Enrichment | Annotation file needed? |
|------------------|------------------|------------------|------------------|
| msigdbr-supported (e.g. Homo sapiens) | Ensembl IDs | MSigDB-based | No (optional, for symbols only) |
| msigdbr-supported (e.g. Homo sapiens) | Not Ensembl IDs | MSigDB-based | Yes (must include an Ensembl ID column) |
| any | any | Annotation-based | Yes, mandatory for enrichment |
| any | any | No GSEA/GESECA needed | No, only DE analysis available |


------------------------------------------------------------------------

# Step 1: Source Genome Annotations

> Add somethign here about this is the same annotation used on the dowstrean RNASeq Aanlysis to create the SE file

## 1.1 Ensembl-based organisms (human, mouse, other msigdbr species)

Example for  `Homo_sapiens_GRCh38` 


```{bash, eval=FALSE}
FILES_DIR="inst/extdata/Ensembl" 
GENOME_NAME="Homo_sapiens.GRCh38" 
VERSION=113

# File used for 
wget -P "$FILES_DIR" \
  "https://ftp.ensembl.org/pub/release-${VERSION}/gtf/homo_sapiens/${GENOME_NAME}.${VERSION}.gtf.gz"

# File used for 
wget -P "$FILES_DIR" \ 
  "https://ftp.ensembl.org/pub/release-${VERSION}/fasta/homo_sapiens/dna/${GENOME_NAME}.dna_sm.primary_assembly.fa.gz"
```

## 1.2 NCBI/GFF3-based organisms (bacteria and others outside Ensembl)

1.  Go to [NCBI Assembly](https://www.ncbi.nlm.nih.gov/assembly/)
2.  Search for your genome accession (e.g., `GCF_000013465.1` for *Staphylococcus aureus* subsp. aureus FPR3757)
3.  Download the assembly in GFF3 format
4.  Unzip and save in your working directory

**Alternative: Use NCBI Datasets CLI**

```{bash, eval = FALSE}
# Install NCBI Datasets CLI if you haven't already. This should be run in your terminal.
conda activate ncbi_datasets 

# Update the genome accession and directory as needed
FILES_DIR="inst/extdata/genomes" 
GENOME_NAME="GCF_000013465.1" 

# Update to download the desired files (genome sequence, GenBank flat file, GFF3...)
datasets download genome accession "$GENOME_NAME" \
            --include genome,gbff,gff3,seq-report \
            --filename "$FILES_DIR/$GENOME_NAME.zip"
            
# Unzip
unzip "$FILES_DIR/$GENOME_NAME.zip" -d "$FILES_DIR"
# Individual files can be renamed to include the genome name for easier identification.
```

## 1.3 Generate your own GFF3 file (Optional)

For genomes not available in NCBI, you can generate your own GFF3 file using Prokka [@Seemann2014] or similar annotation tools. Ensure your GFF3 includes necessary attributes (e.g., `locus_tag`, `product`, etc.).

------------------------------------------------------------------------

# Step 2: Parse Source File into an Annotation Data Frame

## 2.1 From a GTF (Ensembl)

CONTENTO minimal requirement is a `Geneid` column. We can import this from the `gene_id` attribute in the GTF file using `rtracklayer`, along with other useful information such as gene name and biotype.

```{r}
# Set up your R environment and read the GTF file. Adapt based on the GTF name and location
library(tidyverse)
library(here)
library(rtracklayer)

GENOMES_DIR <- here("inst/extdata/Ensembl")
GENOME_NAME <- "Homo_sapiens.GRCh38.113"

# Import GTF file
gtf_data <- import(file.path(GENOMES_DIR, paste0(GENOME_NAME, ".gtf.gz")))

# Extract gene annotations. For available column names check colnames(mcols(gtf_data))
annotations <- data.frame(
  Geneid = mcols(gtf_data)$gene_id,
  gene = mcols(gtf_data)$gene_name,
  biotype = mcols(gtf_data)$gene_biotype
)
# Sanity check: Ensure there are no duplicated Gene IDs
annotations <- annotations[!duplicated(annotations$Geneid), ]

# Save the annotations to a CSV/RDS file
write_csv(annotations, file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.csv")))
saveRDS(annotations, file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.rds")))
```

## 2.2 From a GFF3 (NCBI/Prokka)

CONTENTO minimal requirement is a `Geneid` column. The GFF3 file contains this information as `locus_tag` which we can parse along with other useful information from the `attributes` column.

```{r}
# Set up your R environment and read the GFF3 file. Adapt based on the GFF3 name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

gff_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, ".gff"))

# Read the GFF3 file into a data frame
gff_df <- read_delim(
  gff_file,
  delim = "\t",
  col_names = c("seqname", "source", "feature", "start", "end", "score", "strand", "frame", "attributes"),
  comment = "#",
  show_col_types = FALSE
)

# Filter for CDS (protein-coding genes)
gff_cds <- gff_df |>
  filter(feature == "CDS")

# Function to parse the attributes column of GFF3
parse_gff3_attributes <- function(attributes_string) {
  pairs <- str_split(attributes_string, ";")[[1]]
  
  attr_list <- list()
  
  for (pair in pairs) {
    if (pair == "") next
    
    parts <- str_split(pair, "=", n = 2)[[1]]
    
    if (length(parts) == 2) {
      key <- parts[1]
      value <- parts[2]
      attr_list[[key]] <- value
    }
  }
  
  return(attr_list)
}

# Apply the parsing function to the attributes column and create new columns for each key
annotations <- gff_cds |>
  rowwise() |>
  mutate(attr_list = list(parse_gff3_attributes(attributes)))  |>
  ungroup() |>
  unnest_wider(attr_list, names_repair = "universal") |> # Unnest all key=value pairs as columns
  select(-attributes) |>
  rename(Geneid = locus_tag) # Required column for CONTENTO

# Replace commas in GO-related annotation fields with "!!!" to be used in multiple groupings in functional analysis
annotations <- annotations |>
  mutate(across(c(Ontology_term, go_function, go_process, go_component), ~ gsub(",", "!!!", .x)))

# Aggregate CDS fragments into gene-level annotations (handles programmed frameshifts)
annotations <- annotations |>
  group_by(Geneid) |>
  summarise(
    start  = min(start),
    end    = max(end),
    strand = unique(strand),   # should be invariant per gene
    across(
      -c(start, end, strand, frame),
      first
    ),
    .groups = "drop"
  )

# Sanity check: Ensure there are no duplicated Gene IDs
sum(duplicated(annotations$Geneid))

# Save the annotations to a CSV/RDS file
write_csv(annotations, file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.csv")))
saveRDS(annotations, file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.rds")))
```

------------------------------------------------------------------------

# Step 3: Add Additional Annotations (Optional)

The GTF/GFF3 derived file provides basic gene information, but you can significantly enhance your annotations by running additional bioinformatic tools.

These annotations are useful in CONTENTO for:

- **Functional enrichment analysis (GSEA)** using for example COG and KEGG categories
- **Gene exploration** - displaying comprehensive functional information for individual genes

## 3.1 Merge Custom Annotations

You can merge annotations from any source that contains gene identifiers (Gene IDs). These can be sourced from:

- Public repositories
- Your own experimental data (e.g., Tn-seq, RNA-seq, proteomics)
- Manually curated annotations from literature searches

**In this example**, we demonstrate using RegPrecise regulon annotations, but the same approach works for any custom annotation file.

::: callout-tip
### Found a Useful Annotation Source?

Open an [issue on GitHub](https://github.com/KLemonLab/RNASeqApp/issues) to share it with the community! Include the name, URL, and how you used it.
:::

### Requirements and Format

**Key requirement:** Your custom annotation file must have a column with gene identifiers that match the `Geneid` column in your GTF/GFF3 derived annotations. This column might have a different name in your custom file (e.g., "Geneid", "Gene ID", "Locus tag", ...), but it should contain the same values as `Geneid`.

**Important notes:**

- **Columns:** Your custom annotation file can have as many additional columns as you want. All columns will be merged into the final dataframe, or you can select only a subset of columns before merging. In the merging step, you can **add a prefix** to custom annotation columns (e.g., `RegPrecise_`, `reg_`) to avoid name clashes and easily identify annotation sources.

- **Gene coverage:** No requirement for the same number of genes as in the GTF/GFF3 derived annotations, as long as there is a common key (`Geneid`) for merging. Genes in GTF/GFF3 derived annotations without a corresponding custom annotation entry will be included in the final dataframe with `NA` values for the custom columns.

- **Multiple annotations per gene:** There should be only **one entry per gene** in the custom annotation file. If a gene has multiple annotations, concatenate them into a single string separated by !!! (e.g., `"regulon1!!!regulon2"`). See the anvi'o example below for how to concatenate multiple values. CONTENTO will parse these and count the gene as belonging to each category for functional enrichment analysis.

```{r}
# Set up your R environment and read the anvi'o annotation file. Adapt based on the file name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

# Load your GFF3-derived annotations
annotations <- read_csv(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.csv")))

# Load custom annotation file tab delimited
custom_annot <- read_delim(file.path(GENOMES_DIR, "USA300_FPR3757_RegPrecise.txt"))

# Merge the custom annotations with the GFF3-derived annotations using locus_tag as the key
annotations_merged <- annotations |>
  left_join(custom_annot, by = c("Geneid" = "locus_tag")) |> # Adjust to the "Geneid" column name in your custom annotation file
  rename_with(
    ~ paste0("RegPrecise_", .), # Prefix custom annotation columns to avoid name clashes with GFF3-derived annotations
    -all_of(colnames(annotations))  
  ) 

# Save the annotations to a CSV/RDS file
output_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise.csv"))
write_csv(annotations_merged, output_file)
saveRDS(annotations_merged, file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise.rds"))) 
```

## 3.2 Enhance Annotations with Anvi'o (Bacterial Genomes)

**Anvi'o** is a comprehensive platform that integrates multiple annotation sources and can be used to generate a rich annotation file for bacterial genomes. It can be run on any genome in GenBank format and will produce a single annotation file with multiple functional annotations for each gene.

- **HMMs** - Generic protein domain detection
- **COG24** - Clusters of Orthologous Groups (excellent for functional enrichment)
- **KEGG** - Kyoto Encyclopedia of Genes and Genomes pathways (excellent for functional enrichment)
- **Pfams** - Protein family annotations
- **CAZymes** - Carbohydrate-active enzyme classification

### Run Anvi'o Annotation Pipeline

For more details on how to run anvi'o see the [Anvi'o documentation](https://anvio.org/learn/).

```{bash, eval = FALSE}
# Install Anvio if you haven't already. This should be run in your terminal.
conda activate anvio-9

# Update the genome accession and directory as needed
FILES_DIR="inst/extdata/genomes" 
GENOME_NAME="GCF_000013465.1" 

# Process the GenBank file to extract gene calls, functions, and fasta sequences 
anvi-script-process-genbank -i "$FILES_DIR/$GENOME_NAME.gbff" \
--output-gene-calls "$FILES_DIR/${GENOME_NAME}_calls.txt" \
--output-functions "$FILES_DIR/${GENOME_NAME}_prodigal.txt" \
--output-fasta "$FILES_DIR/${GENOME_NAME}.fa" \
--annotation-source prodigal \
--include-locus-tags-as-functions

# Create anvi'o contigs database from the fasta file and gene calls
anvi-gen-contigs-database -f "$FILES_DIR/${GENOME_NAME}.fa" \
-o "$FILES_DIR/${GENOME_NAME}_anvio.db" \
--external-gene-calls "$FILES_DIR/${GENOME_NAME}_calls.txt" \
-n "$GENOME_NAME"

# Import functions from prodigal output
anvi-import-functions -c "$FILES_DIR/${GENOME_NAME}_anvio.db" \
-i "$FILES_DIR/${GENOME_NAME}_prodigal.txt"

# Run HMM annotations 
anvi-run-hmms -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run NCBI COG annotations with DIAMOND
anvi-run-ncbi-cogs -T 8 --search-with diamond -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run KEGG KOfam annotations
anvi-run-kegg-kofams -T 8 -E 1e-15 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run Pfam domain annotations
anvi-run-pfams -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run CAZyme annotations
anvi-run-cazymes -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Export all functions to text file
anvi-export-functions -c "$FILES_DIR/${GENOME_NAME}_anvio.db" \
-o "$FILES_DIR/${GENOME_NAME}_annot.txt"

# Export contigs stats and database info
anvi-display-contigs-stats "$FILES_DIR/${GENOME_NAME}_anvio.db" \
--report-as-text -o "$FILES_DIR/${GENOME_NAME}_stats.txt"

anvi-db-info "$FILES_DIR/${GENOME_NAME}_anvio.db" &>> "$FILES_DIR/${GENOME_NAME}_log.txt"
```

### Parse Anvi'o Annotations into CSV

The resulting anvi'o annotation file contains comprehensive functional annotations for each gene, which can be parsed into a CSV format with the following R code.

```{r}
# Set up your R environment and read the anvi'o annotation file. Adapt based on the file name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

# Read the anvi'o annotation 
anvio_annot <- read_delim(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.txt")), delim = "\t") |>
  mutate(source = as.factor(source)) |>
  mutate(across(where(is.character), trimws)) |>
  select(-e_value) |> # Exclude e-values to keep dataframe simple
  rename("acc" = "accession", "func" = "function") |> # Shorter names
  mutate(across(c(acc, func), trimws))

# Convert from long format (multiple rows per gene) to wide format (one row per gene)
anvio_annot <- anvio_annot |>
  pivot_wider(
    names_from = source,
    values_from = c(acc, func),
    values_fn = list(
      acc = ~ paste(unique(.), collapse = "!!!"), 
      func = ~ paste(unique(.), collapse = "!!!"))
  ) 
```

### Parse KEGG Hierarchical Annotations

The KEGG annotations from anvi'o come as hierarchical levels separated by semicolons (e.g., "Metabolism; Carbohydrate metabolism; Glycolysis / Gluconeogenesis"). To facilitate functional enrichment analysis and gene exploration in CONTENTO, it's useful to split these into separate columns for each level of the KEGG hierarchy (e.g., KEGG Group, KEGG Pathway).

```{r}
# Split multiple KEGG annotations into separate rows (one row per KEGG annotation) 
anvio_annot <- anvio_annot |>
  separate_rows(func_KEGG_Class, sep = ",")

# Split into hierarchy levels (separated by ;)
anvio_annot <- anvio_annot |>
  separate(func_KEGG_Class, 
           into = c("func_KEGG_Class", "func_KEGG_Subclass", "func_KEGG_Pathway"),
           sep = ";",
           extra = "drop",
           fill = "right")

# Collapse back to one row per gene
anvio_annot <- anvio_annot |>
  group_by(across(-starts_with("func_KEGG_"))) |>
  summarise(
    func_KEGG_Class = paste(unique(na.omit(func_KEGG_Class)), collapse = "!!!"),
    func_KEGG_Subclass = paste(unique(na.omit(func_KEGG_Subclass)), collapse = "!!!"),
    func_KEGG_Pathway = paste(unique(na.omit(func_KEGG_Pathway)), collapse = "!!!"),
    .groups = "drop"
  ) |>
  mutate(across(starts_with("func_KEGG_"), ~ na_if(., "")))
```

### Add AA Sequences to Annotations (Optional)

You can also merge the anvi'o annotations with the aa sequences from the calls file to have a comprehensive annotation dataframe with all the information for each gene. This can be useful for gene exploration in CONTENTO, where you can display the sequence along with the functional annotations for each gene.

```{r}
# Read calls file to get aa sequences
anvio_calls <- read_delim(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_calls.txt")), delim = "\t") |>
  select(gene_callers_id, aa_sequence)

# Merge with anvio_annot by gene_callers_id
anvio_annot <- anvio_annot |>
  left_join(anvio_calls, by = "gene_callers_id")
```

### Merge Anvi'o Annotations 

Finally, you should merge the anvi'o annotations with the GFF3-derived annotations and the custom annotations from RegPrecise to create a comprehensive annotation CSV file for CONTENTO.

```{r}
# Load GFF3-derived annotations (with RegPrecise already merged, or use original if skipping section 3.1)
annotations <- read_csv(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise.csv")))

# Merge anvi'o annotations using GENBANK_LOCUS_TAG (the column added by anvi'o) as a key for merging, which should match the Geneid column in the GFF3-derived annotations. 
annotations_merged <- annotations |>
  left_join(anvio_annot, by = c("Geneid" = "func_GENBANK_LOCUS_TAG")) |> 
  rename_with(
    ~ paste0("anvio_", .), # Prefix anvio annotation columns 
    -all_of(colnames(annotations))  
  ) 

# Save the annotations to a CSV/RDS file
output_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise_Anvio.csv"))
write_csv(annotations_merged, output_file)
saveRDS(annotations_merged, file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise_Anvio.rds")))
```


------------------------------------------------------------------------

# Step 4: Verify Your Annotation File

Before uploading, confirm:

- [ ] `Geneid` column exists, is character type, and matches `rownames(SE)` exactly (same values, same case).
- [ ] Functional columns (if any) use `!!!` to separate multiple values per gene — required for GSEA/GESECA to parse multi-membership correctly.
- [ ] If you want MSigDB and your rownames aren't Ensembl-style, at least one column contains Ensembl-style IDs (`ENSG...`, `ENSMUSG...`, etc.) — CONTENTO auto-detects this column.
- [ ] For genomic neighborhood analysis (bacteria), `start`, `end`, and `strand` columns are present and numeric/character as appropriate.
- [ ] File is saved as `.rds` containing a data frame (not a tibble-only object issue, not a list).

If a feature is greyed out or missing in the app (e.g. MSigDB radio button absent, neighborhood tab missing), it's almost always one of the above checks failing silently.




