---
title: "Create Bacterial Genome Annotations"
author: "Isabel Fernandez Escapa"
date: last-modified
execute:
  eval: false
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
---

# Overview

This tutorial shows how to create a functional annotation CSV file with gene IDs, descriptions, and annotations to be uploaded into the CONTENTO App.

------------------------------------------------------------------------

# Step 1: Source Genome Annotations

## 1.1 Get Your Genome from NCBI

1.  Go to [NCBI Assembly](https://www.ncbi.nlm.nih.gov/assembly/)
2.  Search for your genome accession (e.g., `GCF_000013465.1` for *Staphylococcus aureus* subsp. aureus FPR3757)
3.  Click on the result and downoload the assembly in GFF3 format.
4.  Unzip the file and save it in your working directory.

**Alternative: Use NCBI Datasets CLI**

```{bash, eval = FALSE}
# Install NCBI Datasets CLI if you haven't already. This should be run in your terminal.
conda activate ncbi_datasets 

# Update the genome accession and directory as needed
FILES_DIR="inst/extdata/genomes" 
GENOME_NAME="GCF_000013465.1" 

# Update to download the desired files (genome sequence, GenBank flat file, GFF3...)
datasets download genome accession "$GENOME_NAME" \
            --include genome,gbff,gff3,seq-report \
            --filename "$FILES_DIR/$GENOME_NAME.zip"
            
# Unzip
unzip "$FILES_DIR/$GENOME_NAME.zip" -d "$FILES_DIR"
# Individual files can be renamed to include the genome name for easier identification.
```

## 1.2 Generate your own GFF3 file (Optional)

For genomes not available in NCBI, you can generate your own GFF3 file using Prokka [@Seemann2014] or similar annotation tools. Just make sure to include the necessary attributes (e.g., `locus_tag`, `product`, ...) in the output GFF3.

------------------------------------------------------------------------

# Step 2: Parse Annotation into CSV

CONTENTO requires a CSV file with gene IDs, descriptions, and annotations. The GFF3 file contains this information in the `attributes` column, which we can parse to create a structured annotation CSV.

```{r}
# Set up your R environment and read the GFF3 file. Adapt based on the GFF3 name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

gff_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, ".gff"))

# Read the GFF3 file into a data frame
gff_df <- read_delim(
  gff_file,
  delim = "\t",
  col_names = c("seqname", "source", "feature", "start", "end", "score", "strand", "frame", "attributes"),
  comment = "#",
  show_col_types = FALSE
)

# Filter for CDS (protein-coding genes)
gff_cds <- gff_df |>
  filter(feature == "CDS")

# Function to parse the attributes column of GFF3
parse_gff3_attributes <- function(attributes_string) {
  pairs <- str_split(attributes_string, ";")[[1]]
  
  attr_list <- list()
  
  for (pair in pairs) {
    if (pair == "") next
    
    parts <- str_split(pair, "=", n = 2)[[1]]
    
    if (length(parts) == 2) {
      key <- parts[1]
      value <- parts[2]
      attr_list[[key]] <- value
    }
  }
  
  return(attr_list)
}

# Apply the parsing function to the attributes column and create new columns for each key
annotations <- gff_cds |>
  rowwise() |>
  mutate(attr_list = list(parse_gff3_attributes(attributes)))  |>
  ungroup() |>
  unnest_wider(attr_list, names_repair = "universal") |> # Unnest all key=value pairs as columns
  select(-attributes) # Optionally remove the original attributes column to keep the dataframe clean

# Save the annotations to a CSV file
output_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.csv"))
write_csv(annotations, output_file)
```

# Step 3: Add Additional Annotations (Optional)

The GFF3 file provides basic gene information, but you can significantly enhance your annotations by running additional bioinformatic tools.

These annotations are useful in CONTENTO for:

- **Functional enrichment analysis (GSEA)** using for example COG and KEGG categories
- **Gene exploration** - displaying comprehensive functional information for individual genes

## 3.1 Merge Custom Annotations

You can merge annotations from any source that contains gene identifiers (locus tags). These can be sourced from:

- Public repositories
- Your own experimental data (e.g., Tn-seq, RNA-seq, proteomics)
- Manually curated annotations from literature searches

**In this example**, we demonstrate using RegPrecise regulon annotations, but the same approach works for any custom annotation file.

::: callout-tip
## Found a Useful Annotation Source?

Open an [issue on GitHub](https://github.com/KLemonLab/RNASeqApp/issues) to share it with the community! Include the name, URL, and how you used it.
:::

### Requirements and Format

**Key requirement:** Your custom annotation file must have a column with gene identifiers that match the `locus_tag` column in your GFF3-derived annotations. This column might have a different name in your custom file (e.g., "Geneid", "Gene ID", "Locus tag", ...), but it should contain the same values as `locus_tag`.

**Important notes:**

- **Columns:** Your custom annotation file can have as many additional columns as you want. All columns will be merged into the final dataframe, or you can select only a subset of columns before merging. In the merging step, you can **add a prefix** to custom annotation columns (e.g., `RegPrecise_`, `reg_`) to avoid name clashes and easily identify annotation sources.

- **Gene coverage:** No requirement for the same number of genes as in the GFF3-derived annotations, as long as there is a common key (`locus_tag`) for merging. Genes in GFF3-derived annotations without a corresponding custom annotation entry will be included in the final dataframe with `NA` values for the custom columns.

- **Multiple annotations per gene:** There should be only **one entry per gene** in the custom annotation file. If a gene has multiple annotations, concatenate them into a single string separated by commas (e.g., `"regulon1,regulon2"`). See the anvi'o example below for how to concatenate multiple values. CONTENTO will parse these and count the gene as belonging to each category for functional enrichment analysis.

```{r}
# Set up your R environment and read the anvi'o annotation file. Adapt based on the file name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

# Load your GFF3-derived annotations
annotations <- read_csv(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.csv")))

# Load custom annotation file tab delimited
custom_annot <- read_delim(file.path(GENOMES_DIR, "USA300_FPR3757_RegPrecise.txt"))

# Merge the custom annotations with the GFF3-derived annotations using locus_tag as the key
annotations_merged <- annotations |>
  left_join(custom_annot, by = c("locus_tag" = "Geneid")) |> # Adjust to the "locus_tag" column name in your custom annotation file
  rename_with(
    ~ paste0("RegPrecise_", .), # Prefix custom annotation columns to avoid name clashes with GFF3-derived annotations
    -all_of(colnames(annotations))  
  ) 

# Save the annotations to a CSV file
output_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise.csv"))
write_csv(annotations_merged, output_file)
```

## 3.2 Enhance Annotations with Anvi'o

**Anvi'o** is a comprehensive platform that integrates multiple annotation sources:

- **HMMs** - Generic protein domain detection
- **COG24** - Clusters of Orthologous Groups (excellent for functional enrichment)
- **KEGG** - Kyoto Encyclopedia of Genes and Genomes pathways (excellent for functional enrichment)
- **Pfams** - Protein family annotations
- **CAZymes** - Carbohydrate-active enzyme classification

```{bash, eval = FALSE}
# Install Anvio if you haven't already. This should be run in your terminal.
conda activate anvio-9

# Update the genome accession and directory as needed
FILES_DIR="inst/extdata/genomes" 
GENOME_NAME="GCF_000013465.1" 

# Process the GenBank file to extract gene calls, functions, and fasta sequences 
anvi-script-process-genbank -i "$FILES_DIR/$GENOME_NAME.gbff" \
--output-gene-calls "$FILES_DIR/${GENOME_NAME}_calls.txt" \
--output-functions "$FILES_DIR/${GENOME_NAME}_prodigal.txt" \
--output-fasta "$FILES_DIR/${GENOME_NAME}.fa" \
--annotation-source prodigal \
--include-locus-tags-as-functions

# Create anvi'o contigs database from the fasta file and gene calls
anvi-gen-contigs-database -f "$FILES_DIR/${GENOME_NAME}.fa" \
-o "$FILES_DIR/${GENOME_NAME}_anvio.db" \
--external-gene-calls "$FILES_DIR/${GENOME_NAME}_calls.txt" \
-n "$GENOME_NAME"

# Import functions from prodigal output
anvi-import-functions -c "$FILES_DIR/${GENOME_NAME}_anvio.db" \
-i "$FILES_DIR/${GENOME_NAME}_prodigal.txt"

# Run HMM annotations 
anvi-run-hmms -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run NCBI COG annotations with DIAMOND
anvi-run-ncbi-cogs -T 8 --search-with diamond -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run KEGG KOfam annotations
anvi-run-kegg-kofams -T 8 -E 1e-15 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run Pfam domain annotations
anvi-run-pfams -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Run CAZyme annotations
anvi-run-cazymes -T 8 -c "$FILES_DIR/${GENOME_NAME}_anvio.db"

# Export all functions to text file
anvi-export-functions -c "$FILES_DIR/${GENOME_NAME}_anvio.db" \
-o "$FILES_DIR/${GENOME_NAME}_annot.txt"

# Export contigs stats and database info
anvi-display-contigs-stats "$FILES_DIR/${GENOME_NAME}_anvio.db" \
--report-as-text -o "$FILES_DIR/${GENOME_NAME}_stats.txt"

anvi-db-info "$FILES_DIR/${GENOME_NAME}_anvio.db" &>> "$FILES_DIR/${GENOME_NAME}_log.txt"
```

The resulting anvi'o annotation file contains comprehensive functional annotations for each gene, which can be parsed into a CSV format with the following R code.

```{r}
# Set up your R environment and read the anvi'o annotation file. Adapt based on the file name and location
library(tidyverse)
library(here)

GENOMES_DIR <- here("inst/extdata/genomes")
GENOME_NAME <- "GCF_000013465.1"

# Read the anvi'o annotation 
anvio_annot <- read_delim(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot.txt")), delim = "\t") |>
  mutate(source = as.factor(source)) |>
  mutate(across(where(is.character), trimws)) |>
  select(-e_value) |> # Exclude e-values to keep dataframe simple
  rename("acc" = "accession", "func" = "function") # Shorter names

# Convert from long format (multiple rows per gene) to wide format (one row per gene)
anvio_annot <- anvio_annot |>
  pivot_wider(
    names_from = source,
    values_from = c(acc, func),
    values_fn = list(
      acc = ~ paste(unique(.), collapse = ","), 
      func = ~ paste(unique(.), collapse = ","))
  ) 
```

You can also merge the anvi'o annotations with the aa sequences from the calls file to have a comprehensive annotation dataframe with all the information for each gene. This can be useful for gene exploration in CONTENTO, where you can display the sequence along with the functional annotations for each gene.

```{r}
# Read calls file to get aa sequences
anvio_calls <- read_delim(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_calls.txt")), delim = "\t") |>
  select(gene_callers_id, aa_sequence)

# Merge with anvio_annot by gene_callers_id
anvio_annot <- anvio_annot |>
  left_join(anvio_calls, by = "gene_callers_id")
```

Finally, you can merge the anvi'o annotations with the GFF3-derived annotations and the custom annotations from RegPrecise to create a comprehensive annotation CSV file for CONTENTO.

```{r}
# Load GFF3-derived annotations (with RegPrecise already merged, or use original if skipping section 3.1)
annotations <- read_csv(file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise.csv")))

# Merge anvi'o annotations using GENBANK_LOCUS_TAG (the column added by anvi'o) as a key for merging, which should match the locus_tag column in the GFF3-derived annotations. 
annotations_merged <- annotations |>
  left_join(anvio_annot, by = c("locus_tag" = "func_GENBANK_LOCUS_TAG")) |> 
  rename_with(
    ~ paste0("anvio_", .), # Prefix anvio annotation columns 
    -all_of(colnames(annotations))  
  ) 

# Save the annotations to a CSV file
output_file <- file.path(GENOMES_DIR, paste0(GENOME_NAME, "_annot_RegPrecise_Anvio.csv"))
write_csv(annotations_merged, output_file)
```
