---
title: "Create SE for CONTENTO"
date: last-modified
execute:
  eval: false
  echo: true
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
---





The [**tximport**](https://www.bioconductor.org/packages/release/bioc/html/tximport.html) [@tximport-2] function reads in gene-level quantification results from RSEM and organizes them into a structured list for downstream analysis. It extracts estimated counts, transcript lengths, and abundance values, making it easier to integrate with DESeq2. Since some transcripts may have a reported length of 0, we update those to 1 to prevent issues ensuring valid calculations during analysis.

```{r}

```


```{r}
# Select HNO metadata
meta_HNO <- meta %>%
  filter(media == "HNO") %>%
  droplevels()

meta_HNOt8 <- meta_HNO %>% 
  filter(time == "t08") %>%
  droplevels()

saveRDS(meta_HNO, file.path(metadata_folder, "RNASeq_5772_AIB_Human_meta.rds"))
saveRDS(meta_HNOt8, file.path(metadata_folder, "RNASeq_5772_AIB_Human_meta_t8.rds"))

# Import data using tximport
files <- file.path(rsem_folder, paste0(meta_HNO$file_base, ".genes.results"))
names(files) <- meta_HNO$label
txi <- tximport(files, type = "rsem", txIn = FALSE, txOut = FALSE)
txi$length[txi$length == 0] <- 1

files <- file.path(rsem_folder, paste0(meta_HNOt8$file_base, ".genes.results"))
names(files) <- meta_HNOt8$label
txi_t8 <- tximport(files, type = "rsem", txIn = FALSE, txOut = FALSE)
txi_t8$length[txi_t8$length == 0] <- 1

# Now save the RDS files
saveRDS(txi, file = file.path(DESeq_folder, "HumanReads_txi.rds"))
saveRDS(txi_t8, file = file.path(DESeq_folder, "HumanReads_t8_txi.rds"))
```