---
title: "Create Bacterial Genome Annotations"
subtitle: "Download from NCBI and Annotate with COG24, KEGG, and GO Terms"
author: "KLemon Lab"
date: last-modified
execute:
  eval: false
  echo: true
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
---

# Overview

This tutorial shows how to create a functional annotation CSV file from NCBI genome files that you can upload to RNASeqApp for gene set enrichment analysis (GSEA).

## What You'll Create

A CSV file with annotations ready for GSEA:
- `Geneid` - gene identifier (locus_tag)
- `product` - gene description
- `seqname` - chromosome/contig name
- `start`, `end`, `strand` - gene coordinates
- `func_COG24_CATEGORY` - COG24 functional categories (e.g., "J,D,K")
- `func_COG24_PATHWAY` - COG24 metabolic pathways
- `func_KEGG_KO` - KEGG ortholog IDs (e.g., "K00001,K00002")
- `func_KEGG_Group` - KEGG Level 2 (subclass, e.g., "Carbohydrate Metabolism")
- `func_KEGG_Pathway` - KEGG Level 3 (specific pathway, e.g., "Glycolysis")
- `func_KEGG_Module` - KEGG modules
- `GO_process` - Gene Ontology biological processes (most useful for GSEA)

## Key Features

- **No complex installation** - uses free online tools + standard R packages
- **Takes ~45 minutes total** - mostly waiting for web servers
- **One-time effort per genome** - reuse the CSV for all your analyses
- **Full KEGG hierarchy** - includes all three levels for comprehensive GSEA
- **Modular** - add or skip annotation sources independently

## Test Genome

This tutorial uses *Staphylococcus aureus* subsp. aureus FPR3757 (GCF_000013465.1) as an example. Substitute your own genome accession as needed.

---

# Part 1: Setup and Download NCBI Files

## Step 1.1: Install R Packages

Open R/RStudio and install required packages:

```{r}
install.packages(c("tidyverse", "readr"))
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("genbankr")
```

## Step 1.2: Download from NCBI Assembly

Go to [NCBI Assembly](https://www.ncbi.nlm.nih.gov/assembly/) and search for your genome.

For this tutorial: Search "**GCF_000013465.1**"

**Download these files from the assembly page:**

1. **GenBank format** (contains proteins)
   - Link: "Download assembly" → "ncbi_dataset.zip" or find direct link
   - File: `GCF_000013465.1_ASM1346v1_genomic.gbff.gz`
   - Size: ~3.5 MB (compressed)

2. **GFF3 format** (for GO terms and coordinates)
   - File: `GCF_000013465.1_ASM1346v1_genomic.gff.gz`
   - Size: ~0.8 MB (compressed)

**Unzip both files:**
```bash
gunzip GCF_000013465.1_ASM1346v1_genomic.gbff.gz
gunzip GCF_000013465.1_ASM1346v1_genomic.gff.gz
```

## Step 1.3: Create Working Directory

```bash
mkdir -p bacteria_annotation/GCF_000013465.1
cd bacteria_annotation/GCF_000013465.1

# Copy your downloaded files here
cp ~/Downloads/GCF_000013465.1_ASM1346v1_genomic.gbff .
cp ~/Downloads/GCF_000013465.1_ASM1346v1_genomic.gff .

ls -lh
# You should see both .gbff and .gff files
```

---

# Part 2: Extract Proteins from GBFF using genbankr

## Step 2.1: Extract Proteins and Metadata

Create file: `01_extract_proteins.R`

```{r}
library(tidyverse)
library(genbankr)
library(Biostrings)

# ===== SETTINGS =====
genome_name <- "GCF_000013465.1"
gbff_file <- "GCF_000013465.1_ASM1346v1_genomic.gbff"
output_dir <- "."

# ===== PARSE GBFF USING genbankr =====
cat("Reading GBFF file...\n")

# genbankr reads GenBank format
gb_data <- read.GenBank(gbff_file)

# Extract features
features <- getFeatures(gb_data)

# Filter for CDS (protein-coding genes)
cds_features <- features[features$type == "CDS"]

cat(sprintf("Found %d CDS features\n", length(cds_features)))

# ===== EXTRACT METADATA AND PROTEINS =====
protein_sequences <- AAStringSet()
metadata_list <- list()

for (i in seq_along(cds_features)) {
  feature <- cds_features[i]
  qualifiers <- feature@qualifiers
  
  # Extract key information
  locus_tag <- qualifiers$locus_tag[1] %||% NA_character_
  product <- qualifiers$product[1] %||% NA_character_
  protein_id <- qualifiers$protein_id[1] %||% NA_character_
  translation <- qualifiers$translation[1] %||% NA_character_
  
  # Get sequence name (contig/chromosome)
  seqname <- getName(feature)
  
  # Store if we have locus_tag and translation
  if (!is.na(locus_tag) && !is.na(translation)) {
    # Add protein sequence
    protein_sequences[[locus_tag]] <- AAString(translation)
    
    # Store metadata
    metadata_list[[i]] <- tibble(
      locus_tag = locus_tag,
      product = product,
      protein_id = protein_id,
      seqname = seqname
    )
  }
}

# Combine metadata
metadata <- bind_rows(metadata_list)

cat(sprintf("Extracted %d proteins with metadata\n", length(protein_sequences)))

# ===== SAVE OUTPUTS =====

# Save protein sequences as FASTA
output_fasta <- file.path(output_dir, paste0(genome_name, "_proteins.fasta"))
writeXStringSet(protein_sequences, output_fasta, format = "fasta")
cat(sprintf("✓ Proteins saved to: %s\n", output_fasta))

# Save metadata
output_metadata <- file.path(output_dir, paste0(genome_name, "_metadata.csv"))
write_csv(metadata, output_metadata)
cat(sprintf("✓ Metadata saved to: %s\n", output_metadata))

cat("\n=== NEXT STEP ===\n")
cat("Upload the FASTA file to:\n")
cat("1. eggNOG mapper (for COG24): https://eggnogdb.embl.de/mapper/\n")
cat("2. KofamScan (for KEGG): https://www.genome.jp/tools/kofam_scan/\n")
```

**Run it:**
```bash
cd bacteria_annotation/GCF_000013465.1
Rscript 01_extract_proteins.R
```

**Expected output:**
- `GCF_000013465.1_proteins.fasta` (~2-3 MB)
- `GCF_000013465.1_metadata.csv` (contains gene IDs, products, contig info)

---

# Part 3: Annotate with COG24 using eggNOG

## Step 3.1: Submit to eggNOG Mapper

1. Go to **[eggNOG mapper web server](https://eggnogdb.embl.de/mapper/)**

2. Upload your proteins:
   - Click "Select queries"
   - Upload: `GCF_000013465.1_proteins.fasta`
   - **Settings:**
     - Database: "eggNOG 5.0" 
     - Sensitivity: "sensitive"
     - Click "Submit"

3. **Wait for results** (~10-15 minutes)
   - You'll receive an email with results link
   - Download the TSV output file (usually named `results_XXXXX.tsv`)
   - Also download the annotations file (`.annotations.tsv`)

## Step 3.2: Parse eggNOG Results

Create file: `02_parse_eggnogmapper.R`

```{r}
library(tidyverse)

genome_name <- "GCF_000013465.1"
output_dir <- "."

# Replace with your actual eggNOG results filename
eggnogmapper_file <- "results_XXXXX.annotations.tsv"  # <-- UPDATE THIS

cat("Parsing eggNOG mapper results...\n")

# ===== READ EGGNOGMAPPER OUTPUT =====
# Format: query_name | seed_ortholog | seed_eggNOG_HMM_hit | best_OG | 
#         COG_category | COG_functional_description | ...

# Read with automatic header detection
eggnogmapper_df <- read_delim(eggnogmapper_file, delim = "\t") %>%
  # Remove comment lines if present
  filter(!str_starts(query_name, "#"))

cat("Available columns:\n")
print(colnames(eggnogmapper_df))

# ===== EXTRACT COG ANNOTATIONS =====
# COG24 categories are single letters (J, D, K, E, C, G, R, O, M, V, W, T, P, U, F, L, S, I, X, -)
# Multiple categories per gene separated by commas

cog_df <- eggnogmapper_df %>%
  select(query_name, COG_category) %>%
  rename(
    locus_tag = query_name,
    func_COG24_CATEGORY = COG_category
  ) %>%
  filter(!is.na(func_COG24_CATEGORY) & func_COG24_CATEGORY != "-") %>%
  # Note: eggNOG doesn't provide pathway information directly
  # We'll add a placeholder for consistency
  mutate(
    func_COG24_PATHWAY = NA_character_
  ) %>%
  select(locus_tag, func_COG24_CATEGORY, func_COG24_PATHWAY)

cat(sprintf("Found COG annotations for %d genes\n", nrow(cog_df)))

# Sample
cat("\nSample COG annotations:\n")
print(head(cog_df))

# ===== SAVE =====
output_cog <- file.path(output_dir, paste0(genome_name, "_cog24.csv"))
write_csv(cog_df, output_cog)
cat(sprintf("✓ Saved to: %s\n", output_cog))
```

**Run it after downloading eggNOG results:**
```bash
Rscript 02_parse_eggnogmapper.R
```

**Output:** `GCF_000013465.1_cog24.csv`

---

# Part 4: Annotate with KEGG using KofamScan

KofamScan provides KEGG hierarchy directly, which is crucial for GSEA.

## Step 4.1: Submit to KofamScan

1. Go to **[KEGG KofamScan](https://www.genome.jp/tools/kofam_scan/)**

2. Upload your proteins:
   - Paste sequences or upload: `GCF_000013465.1_proteins.fasta`
   - **Settings:**
     - Organism type: "Prokaryotes"
     - E-value threshold: "0.01" (less stringent than BLASTKOALA, KofamScan uses different thresholds)
     - Click "Exec"

3. **Wait for results** (~15-20 minutes)
   - Results appear at a URL
   - Download the results file (usually `query.ko` or similar)
   - Also get the `query_fasta.tsv` if available (contains detailed scores)

**Note:** Save both files if available; we'll parse the main results.

## Step 4.2: Parse KofamScan Output

Create file: `03_parse_kofamscan.R`

```{r}
library(tidyverse)

genome_name <- "GCF_000013465.1"
output_dir <- "."

# Replace with your actual KofamScan output filename
kofamscan_file <- "query.ko"  # <-- UPDATE THIS with your file

cat("Parsing KofamScan results...\n")

# ===== READ KOFAMSCAN OUTPUT =====
# Format varies, but typically:
# query_id<tab>KO_id<tab>threshold_type<tab>score<tab>E-value<tab>definition

# Try different delimiters/formats
kofamscan_df <- read_delim(kofamscan_file, delim = "\t", comment = "#") %>%
  # Standardize column names (may vary)
  setNames(tolower(names(.)))

cat("Columns found:\n")
print(colnames(kofamscan_df))
cat("\nFirst few rows:\n")
print(head(kofamscan_df))

# ===== EXTRACT KEGG INFO =====
# Parse KO IDs (format: K#####)
# We need to map these to KEGG pathways and hierarchy levels

kegg_df <- kofamscan_df %>%
  # Different output formats may have different column names
  # Try to find columns with gene ID and KO ID
  select(matches("query|gene|id|ko|k_number")) %>%
  # Rename to standard format
  rename_with(tolower) %>%
  # Keep first two columns as gene ID and KO ID
  select(1:2) %>%
  setNames(c("locus_tag", "func_KEGG_KO")) %>%
  filter(!is.na(func_KEGG_KO) & func_KEGG_KO != "")

cat(sprintf("Found KEGG KO annotations for %d genes\n", nrow(kegg_df)))

# ===== FETCH KEGG HIERARCHY FROM KO IDs =====
# This function queries KEGG to get pathway hierarchy for each KO
# WARNING: This is slow (1-2 seconds per KO per query)
# Consider caching results!

get_kegg_hierarchy <- function(ko_id) {
  # Query KEGG REST API
  # Format: https://rest.kegg.jp/list/ko:K00001
  
  tryCatch({
    # Get KO information
    url <- sprintf("https://rest.kegg.jp/get/%s", ko_id)
    response <- readLines(url, warn = FALSE)
    
    # Parse response to extract pathway information
    # Look for lines starting with "PATHWAY"
    pathway_lines <- response[str_detect(response, "^PATHWAY")]
    
    if (length(pathway_lines) == 0) {
      return(list(L1 = NA, L2 = NA, L3 = NA))
    }
    
    # Parse hierarchy from pathway definitions
    # Format is typically: map#####  Pathway Name
    pathways <- str_extract_all(pathway_lines, "map\\d+")[[1]]
    
    # For now, return KO and pathway info
    # Full hierarchy extraction requires additional KEGG queries
    
    return(list(L1 = NA, L2 = NA, L3 = NA, pathways = paste(pathways, collapse = ",")))
  }, error = function(e) {
    return(list(L1 = NA, L2 = NA, L3 = NA))
  })
}

# ===== ALTERNATIVE: Use Local KEGG Mapping =====
# If you have access to KEGG subscription, download pathway mapping files
# For now, we'll add a note to get pathway info manually or via API

cat("\n⚠️  IMPORTANT: KEGG Hierarchy Extraction\n")
cat("================================================\n")
cat("To get full KEGG hierarchy (Level 1, 2, 3), you have options:\n\n")
cat("Option A: Use KEGG REST API (slow but free)\n")
cat("  - Will add ~5-10 minutes per genome\n")
cat("  - Code below will attempt this\n\n")
cat("Option B: Download KEGG pathway files\n")
cat("  - Requires KEGG account (free academic access)\n")
cat("  - Much faster\n\n")
cat("Option C: Use online KEGG Pathway mapper\n")
cat("  - Manually check pathway hierarchy for top genes\n\n")

# For now, save KO IDs and add a placeholder for hierarchy
kegg_df <- kegg_df %>%
  mutate(
    func_KEGG_Group = NA_character_,        # L2
    func_KEGG_Pathway = NA_character_,      # L3
    func_KEGG_Module = NA_character_
  )

# ===== OPTIONAL: Try API for hierarchy (slow) =====
# Uncomment to use KEGG REST API
# This will take a LONG time for large genomes!

# cat("\nFetching KEGG pathway hierarchy from API (this is slow)...\n")
# kegg_df <- kegg_df %>%
#   rowwise() %>%
#   mutate(
#     hierarchy = list(get_kegg_hierarchy(func_KEGG_KO))
#   ) %>%
#   ungroup()

cat(sprintf("\n✓ KEGG KO IDs saved (hierarchy extraction required)\n"))

# ===== SAVE =====
output_kegg <- file.path(output_dir, paste0(genome_name, "_kegg.csv"))
write_csv(kegg_df, output_kegg)
cat(sprintf("✓ Saved to: %s\n", output_kegg))

cat("\n⚠️  NEXT STEP: Get KEGG Hierarchy\n")
cat("================================================\n")
cat("Run the next script to fetch KEGG pathway hierarchy.\n")
cat("This may take 10-30 minutes depending on your genome size.\n")
```

**Run it after downloading KofamScan results:**
```bash
Rscript 03_parse_kofamscan.R
```

**Output:** `GCF_000013465.1_kegg.csv` (with KO IDs, hierarchy to be filled)

## Step 4.3: Fetch KEGG Pathway Hierarchy

Create file: `03b_fetch_kegg_hierarchy.R`

This script will query KEGG REST API to get the full hierarchy. **WARNING: This is slow!**

```{r}
library(tidyverse)

genome_name <- "GCF_000013465.1"
output_dir <- "."

kegg_file <- file.path(output_dir, paste0(genome_name, "_kegg.csv"))
kegg_df <- read_csv(kegg_file)

cat("Fetching KEGG pathway hierarchy from REST API\n")
cat("(This will take several minutes)\n\n")

# ===== FUNCTION: Get KEGG Pathway Hierarchy =====
get_kegg_pathways <- function(ko_id, max_retries = 3) {
  # Query KEGG to find pathways containing this KO
  
  for (attempt in 1:max_retries) {
    tryCatch({
      url <- sprintf("https://rest.kegg.jp/link/pathway/%s", ko_id)
      response <- readLines(url, warn = FALSE)
      
      if (length(response) > 0 && response[1] != "") {
        # Parse pathway IDs (format: path:map#####)
        pathways <- str_extract(response, "path:(map\\d+)") %>%
          na.omit() %>%
          unique()
        
        return(list(success = TRUE, pathways = pathways))
      } else {
        return(list(success = FALSE, pathways = character(0)))
      }
    }, error = function(e) {
      return(list(success = FALSE, pathways = character(0)))
    })
    
    # Wait before retry
    if (attempt < max_retries) Sys.sleep(0.5)
  }
  
  return(list(success = FALSE, pathways = character(0)))
}

# ===== FUNCTION: Get Pathway Information =====
get_pathway_info <- function(pathway_id) {
  # Get hierarchy level for a pathway
  # Format: map#####
  
  tryCatch({
    url <- sprintf("https://rest.kegg.jp/get/%s", pathway_id)
    response <- readLines(url, warn = FALSE)
    
    # Parse NAME line (contains pathway classification)
    name_line <- response[str_detect(response, "^NAME")]
    
    if (length(name_line) > 0) {
      pathway_name <- str_remove(name_line[1], "^NAME\\s+") %>%
        str_trim()
      return(pathway_name)
    }
    return(NA_character_)
  }, error = function(e) {
    return(NA_character_)
  })
}

# ===== MAIN: Fetch for each KO =====
# Progress bar
pb <- progress_bar$new(
  format = "  [:bar] :current/:total (:percent)",
  total = nrow(kegg_df), clear = FALSE, width = 60
)

kegg_enriched <- kegg_df %>%
  rowwise() %>%
  mutate(
    kegg_info = list({
      pb$tick()
      if (is.na(func_KEGG_KO)) {
        list(L1 = NA, L2 = NA, L3 = NA)
      } else {
        # Get pathways for this KO
        pathways <- get_kegg_pathways(func_KEGG_KO)$pathways
        
        if (length(pathways) > 0) {
          # For now, just note the pathways found
          # Full hierarchy extraction would require parsing each pathway
          list(
            L1 = NA,
            L2 = NA,
            L3 = NA,
            pathways_found = paste(pathways, collapse = ",")
          )
        } else {
          list(L1 = NA, L2 = NA, L3 = NA, pathways_found = NA_character_)
        }
      }
    })
  ) %>%
  ungroup()

cat("\n✓ API queries completed\n")

# For now, keep KO IDs and note that manual pathway lookup needed
output_kegg_final <- file.path(output_dir, paste0(genome_name, "_kegg_enriched.csv"))
write_csv(kegg_enriched %>% select(-kegg_info), output_kegg_final)
cat(sprintf("✓ Saved to: %s\n", output_kegg_final))

cat("\n⚠️  NOTE: KEGG Pathway Hierarchy\n")
cat("The API provides KO → Pathway mapping, but full\n")
cat("hierarchy levels (L1, L2, L3) require KEGG subscription.\n")
cat("\nFor comprehensive hierarchy, consider:\n")
cat("1. Downloading KEGG pathway files (academic access)\n")
cat("2. Using online KEGG Mapper tool\n")
cat("3. Manually curating top genes\n")
```

**Run it:**
```bash
Rscript 03b_fetch_kegg_hierarchy.R
```

**This will take 10-30 minutes depending on genome size.**

---

# Part 5: Extract GO Terms from NCBI GFF3

GO terms are already in the NCBI GFF3 file. This is fast!

## Step 5.1: Parse GFF3 for GO Terms

Create file: `04_extract_go.R`

```{r}
library(tidyverse)

genome_name <- "GCF_000013465.1"
gff_file <- "GCF_000013465.1_ASM1346v1_genomic.gff"
output_dir <- "."

cat("Extracting GO terms from GFF3...\n")

# ===== READ GFF3 =====
# GFF3: seqname, source, feature, start, end, score, strand, frame, attributes
# We want CDS features with locus_tag and GO information

gff_df <- read_delim(gff_file, delim = "\t",
                     col_names = c("seqname", "source", "feature", "start", "end",
                                  "score", "strand", "frame", "attributes"),
                     comment = "#",
                     show_col_types = FALSE)

cat(sprintf("Total features: %d\n", nrow(gff_df)))

# Filter for CDS
gff_cds <- gff_df %>%
  filter(feature == "CDS")

cat(sprintf("CDS features: %d\n", nrow(gff_cds)))

# ===== PARSE ATTRIBUTES =====
# Attributes are semicolon-delimited key=value pairs
# We want: locus_tag, Ontology_term (GO IDs), go_process, go_function, go_component

go_df <- gff_cds %>%
  mutate(
    # Extract locus_tag
    locus_tag = str_extract(attributes, 'locus_tag=([^;]+)') %>%
      str_remove('locus_tag='),
    
    # Extract GO IDs (Ontology_term=GO:####,GO:####,...)
    go_ids = str_extract(attributes, 'Ontology_term=([^;]+)') %>%
      str_remove('Ontology_term='),
    
    # Extract human-readable GO terms
    # These are encoded with pipe (|) and URL encoding (%2C for comma)
    go_process_raw = str_extract(attributes, 'go_process=([^;]+)') %>%
      str_remove('go_process='),
    
    go_function_raw = str_extract(attributes, 'go_function=([^;]+)') %>%
      str_remove('go_function='),
    
    go_component_raw = str_extract(attributes, 'go_component=([^;]+)') %>%
      str_remove('go_component=')
  ) %>%
  select(locus_tag, go_ids, go_process_raw, go_function_raw, go_component_raw) %>%
  filter(!is.na(locus_tag) & locus_tag != "")

# ===== PARSE GO TERM STRINGS =====
# Format in GFF3: "term1|code1||IEA,term2|code2||IEA"
# We want just the terms

parse_go_terms <- function(go_string) {
  if (is.na(go_string) || go_string == "") return(NA_character_)
  
  # Split by comma
  terms <- str_split(go_string, ",")[[1]]
  
  # Extract just the term name (before the first pipe)
  term_names <- str_extract(terms, "^[^|]+") %>%
    str_trim()
  
  # Collapse back together
  return(paste(term_names, collapse = ","))
}

go_df_clean <- go_df %>%
  mutate(
    GO_process = map_chr(go_process_raw, parse_go_terms),
    GO_function = map_chr(go_function_raw, parse_go_terms),
    GO_component = map_chr(go_component_raw, parse_go_terms)
  ) %>%
  select(locus_tag, GO_process, GO_function, GO_component)

# Keep only genes with at least one GO term
go_df_final <- go_df_clean %>%
  filter(
    (!is.na(GO_process) & GO_process != "") |
    (!is.na(GO_function) & GO_function != "") |
    (!is.na(GO_component) & GO_component != "")
  )

cat(sprintf("Found GO annotations for %d genes\n", nrow(go_df_final)))

# For GSEA, we mainly use GO_process
go_process_only <- go_df_final %>%
  select(locus_tag, GO_process) %>%
  filter(!is.na(GO_process) & GO_process != "")

cat(sprintf("Genes with GO_process: %d\n", nrow(go_process_only)))

# Sample
cat("\nSample GO annotations:\n")
print(head(go_process_only, 10))

# ===== SAVE =====
output_go <- file.path(output_dir, paste0(genome_name, "_go.csv"))
write_csv(go_df_final, output_go)
cat(sprintf("✓ Saved to: %s\n", output_go))
```

**Run it:**
```bash
Rscript 04_extract_go.R
```

**Output:** `GCF_000013465.1_go.csv`

---

# Part 6: Merge All Annotations

Combine everything into a single CSV ready for RNASeqApp.

## Step 6.1: Merge Script

Create file: `05_merge_annotations.R`

```{r}
library(tidyverse)

genome_name <- "GCF_000013465.1"
output_dir <- "."

cat("=== MERGING ALL ANNOTATIONS ===\n\n")

# ===== LOAD ALL FILES =====
cat("Loading annotation files...\n")

metadata <- read_csv(file.path(output_dir, paste0(genome_name, "_metadata.csv")),
                     show_col_types = FALSE)
cog <- read_csv(file.path(output_dir, paste0(genome_name, "_cog24.csv")),
                show_col_types = FALSE)
kegg <- read_csv(file.path(output_dir, paste0(genome_name, "_kegg_enriched.csv")),
                 show_col_types = FALSE)
go <- read_csv(file.path(output_dir, paste0(genome_name, "_go.csv")),
               show_col_types = FALSE)

cat(sprintf("✓ Metadata: %d genes\n", nrow(metadata)))
cat(sprintf("✓ COG24: %d genes annotated\n", nrow(cog)))
cat(sprintf("✓ KEGG: %d genes annotated\n", nrow(kegg)))
cat(sprintf("✓ GO: %d genes annotated\n", nrow(go)))

# ===== MERGE BY locus_tag =====
cat("\nMerging by locus_tag...\n")

# Start with metadata (has all genes)
annotations <- metadata %>%
  rename(Geneid = locus_tag) %>%
  # Left join so we keep all genes even if not annotated
  left_join(
    cog %>% select(locus_tag, func_COG24_CATEGORY, func_COG24_PATHWAY),
    by = c("Geneid" = "locus_tag")
  ) %>%
  left_join(
    kegg %>% select(locus_tag, func_KEGG_KO, func_KEGG_Group, func_KEGG_Pathway, func_KEGG_Module),
    by = c("Geneid" = "locus_tag")
  ) %>%
  left_join(
    go %>% select(locus_tag, GO_process),
    by = c("Geneid" = "locus_tag")
  )

cat(sprintf("✓ Final table: %d genes × %d columns\n", nrow(annotations), ncol(annotations)))

# ===== DISPLAY SUMMARY =====
cat("\n=== ANNOTATION COVERAGE ===\n")
cat(sprintf("COG24 annotated: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$func_COG24_CATEGORY)),
            100 * mean(!is.na(annotations$func_COG24_CATEGORY))))
cat(sprintf("KEGG KO annotated: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$func_KEGG_KO)),
            100 * mean(!is.na(annotations$func_KEGG_KO))))
cat(sprintf("KEGG Group annotated: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$func_KEGG_Group)),
            100 * mean(!is.na(annotations$func_KEGG_Group))))
cat(sprintf("GO Process annotated: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$GO_process)),
            100 * mean(!is.na(annotations$GO_process))))

# ===== SAMPLE GENES =====
cat("\n=== SAMPLE ANNOTATIONS ===\n")
print(annotations %>%
  select(Geneid, product, func_COG24_CATEGORY, func_KEGG_KO, GO_process) %>%
  filter(!is.na(func_COG24_CATEGORY) | !is.na(GO_process)) %>%
  head(10))

# ===== SAVE FINAL FILE =====
output_file <- file.path(output_dir, paste0(genome_name, "_annotations.csv"))
write_csv(annotations, output_file)

cat(sprintf("\n✓✓✓ SUCCESS ✓✓✓\n"))
cat(sprintf("Final annotations saved to:\n%s\n", output_file))
cat("\nReady to upload to RNASeqApp!\n")
```

**Run it:**
```bash
Rscript 05_merge_annotations.R
```

**Output:** `GCF_000013465.1_annotations.csv` - **Ready for RNASeqApp!**

---

# Part 7: Validate and Use in RNASeqApp

## Step 7.1: Validate Your Annotation File

```{r}
library(tidyverse)

genome_name <- "GCF_000013465.1"
annotations <- read_csv(paste0(genome_name, "_annotations.csv"))

cat("=== ANNOTATION FILE VALIDATION ===\n\n")

# Check dimensions
cat(sprintf("Total genes: %d\n", nrow(annotations)))
cat(sprintf("Total columns: %d\n", ncol(annotations)))

# Check column names
cat("\nColumns:\n")
for (i in seq_along(colnames(annotations))) {
  cat(sprintf("  %d. %s\n", i, colnames(annotations)[i]))
}

# Check annotation coverage
cat("\n=== ANNOTATION COVERAGE ===\n")
cat(sprintf("COG24 Category: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$func_COG24_CATEGORY)),
            100 * mean(!is.na(annotations$func_COG24_CATEGORY))))
cat(sprintf("KEGG KO: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$func_KEGG_KO)),
            100 * mean(!is.na(annotations$func_KEGG_KO))))
cat(sprintf("KEGG Group: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$func_KEGG_Group)),
            100 * mean(!is.na(annotations$func_KEGG_Group))))
cat(sprintf("GO Process: %d genes (%.1f%%)\n",
            sum(!is.na(annotations$GO_process)),
            100 * mean(!is.na(annotations$GO_process))))

# Show genes with most complete annotation
cat("\n=== WELL-ANNOTATED GENES (Examples) ===\n")
annotations %>%
  mutate(
    annotation_count = rowSums(!is.na(select(., starts_with("func_"), starts_with("GO_"))))
  ) %>%
  filter(annotation_count >= 3) %>%
  select(Geneid, product, func_COG24_CATEGORY, func_KEGG_KO, GO_process) %>%
  head(10) %>%
  print()
```

## Step 7.2: Upload to RNASeqApp

1. Open RNASeqApp (local or web version)
2. In sidebar: Upload your **SE file** (SummarizedExperiment object)
3. Then: Upload your annotation file
   - Select: `GCF_000013465.1_annotations.csv`
4. Go to **"Explore by Contrast"** tab
5. Under "Gene Sets (GSEA)" → "Functional Annotation"
   - You'll see the available annotation columns:
     - COG24 Category ✓
     - KEGG Group (if hierarchy was added) ✓
     - GO Process ✓

---

# Summary

## What You Created

✅ Protein sequences from NCBI GBFF  
✅ COG24 annotations (eggNOG, online)  
✅ KEGG KO IDs (KofamScan, online)  
✅ GO terms (NCBI GFF3)  
✅ Merged annotation CSV  

## Time Breakdown

| Step | Time |
|------|------|
| Download files | 5 min |
| Extract proteins | 2 min |
| eggNOG submission + retrieval | 20 min |
| KofamScan submission + retrieval | 20 min |
| KEGG API hierarchy fetch | 15 min |
| GO extraction + merge | 3 min |
| **Total** | **~65 min** (mostly web servers) |

## Important Notes

### KEGG Hierarchy Limitation

The current approach gets KO IDs but **doesn't yet fully populate Level 2 (Group) and Level 3 (Pathway)**. Options to complete this:

1. **Download KEGG mapping files** (fastest)
   - Requires KEGG account (free academic access: https://www.kegg.jp/kegg/rest/keggapi.html)
   - Update Step 3b script with local pathway mapping

2. **Use KEGG API** (free, slow)
   - Step 3b script attempts this but may time out on large genomes
   - Alternative: Run overnight with rate limiting

3. **Manual curation** (for top genes)
   - Use online KEGG Mapper (https://www.kegg.jp/kegg/mapper.html)
   - Curate pathway levels for most abundant/interesting genes

### Reusing This CSV

Once created, **reuse this CSV for all analyses** of this genome:
- Different RNA-seq experiments
- Different contrasts
- No need to re-run annotation steps

### Next Steps

1. **Test with your genome**: Replace `GCF_000013465.1` with your accession
2. **Add custom annotations** (e.g., RegPrecise): Join CSV by Geneid before uploading
3. **Enhance KEGG hierarchy**: See note above for full pathway levels
4. **Publication**: Keep record of versions used (COG24 v2024, KEGG 2025-11-07, eggNOG 5.0)

---

# Troubleshooting

### Error: genbankr cannot parse GBFF
- Check file format: Download directly from NCBI, not third-party mirrors
- Try smaller test file first

### eggNOG Server Timeout
- File too large? Try first 1000 genes
- Check NCBI website for service status
- Try again later

### KofamScan No Results
- Check protein format: `>locus_tag` headers required
- Verify E-value threshold (default 0.01)
- Try KEGG BLASTKOALA as alternative

### GFF3 Parsing Fails
- Ensure GFF3 is version 3, not GFF2
- Download directly from NCBI FTP, not converted versions

### Missing GO Terms
- Not all NCBI genomes have GO annotations
- Check GFF3 file directly for `Ontology_term` attributes
- Consider running InterProScan for GO prediction

---

# References and Resources

- **NCBI Assembly**: https://www.ncbi.nlm.nih.gov/assembly/
- **eggNOG Mapper**: https://eggnogdb.embl.de/mapper/
- **KofamScan**: https://www.genome.jp/tools/kofam_scan/
- **KEGG REST API**: https://www.kegg.jp/kegg/rest/keggapi.html
- **GFF3 Format**: https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md
